---
title: "Final Project 2026 — Yacht Hydrodynamic Resistance"
output:
  html_document:
    df_print: paged
---

## IMPORTANT:

read carefully the instruction for the final project in the pdf file

**Project2026_INFO.pdf**

## Yacht hydrodynamic resistance

**Source.** UCI Machine Learning Repository, <https://archive.ics.uci.edu/dataset/243/yacht+hydrodynamics>. Original record: Gerritsma, Onnink and Versluis, *Geometry, resistance and stability of the Delft Systematic Yacht Hull Series*, 1981.

**Data.** The dataset describes 22 different hull forms (sailing yacht), each tested at multiple Froude numbers in a towing tank, producing 308 hull-form $\times$ Froude-number rows.

Variables:

`LCB` : longitudinal position of the center of buoyancy (% length).

`prismatic` : prismatic coefficient (dimensionless).

`L_disp` : length-displacement ratio.

`beam_draught` : beam-draught ratio.

`L_beam` : length-beam ratio.

`Froude` : Froude number (dimensionless, $Fn = v/\sqrt{gL}$), in [0.125, 0.45]. The dominant predictor.

`resistance` : residuary resistance per unit weight of displacement (target variable).

**Task.** The physics of the problem says that for sailing yachts the residuary resistance follows a power law of the Froude number with exponent close to 4 in the bulk regime. A good Bayesian project explores this:

1.  **Log-log regression.** Fit $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude}) + \sum_k \beta_k\,\mathrm{hull}_k + \varepsilon\,.$$ Verify that $\hat\beta_F \approx 4$.

2.  **BAS / BMA.** Use `bas.lm` to enumerate the $2^5$ submodels over the five hull-geometry covariates and identify which of them (LCB, prismatic, L/disp, B/draught, L/beam) carry signal once `Froude` is in the model.

3.  **Correct for heteroscedasticity.** Train the nonlinear model $$\mathrm{resistance} = k\,\mathrm{Froude}^{\gamma_F}\,\exp\left\{\sum_k \beta_k\,\mathrm{hull}_k \right\} + \varepsilon$$ where only the form factors that appear relevant in the model selection phase are kept for the nonlinear modelling. Compare the results with the log-log regression and discuss the impact of the heteroscedasticity correction on the inference on $\gamma_F$ and on the hull-geometry covariates.

4.  Plot the posterior predictive curves and compare the curves with or without the treatment of the heteroscedasticity and compare the curves with the data.

```{r}
rm(list=ls())
# Yacht hydrodynamics — UCI #243
yh = read.csv("data/yacht_hydro.csv")
dim(yh)
head(yh)
summary(yh)

# The power-law in Fn is best seen on log-log axes;
# resistance > 0 everywhere (min = 0.01), so plain log() needs no offset
plot(log(yh$Froude), log(yh$resistance),
     pch = 16, cex = 0.5,
     xlab = "log(Froude)", ylab = "log(resistance)",
     main = "Yacht hydrodynamic resistance — power-law scaling")
abline(lm(log(resistance) ~ log(Froude), data = yh),
       col = "red", lwd = 2)
```

## 1. Log-log regression (Bayesian, reference prior)

We fit the log-log model $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude})
  + \sum_{k=1}^{5}\beta_k\,\mathrm{hull}_k + \varepsilon,
  \qquad \varepsilon \sim \mathcal N(0,\sigma^2),$$ with the non-informative reference (Jeffreys) prior $\pi(\beta\mid\sigma^2)\propto 1$, $\pi(\sigma^2)\propto 1/\sigma^2$. The smallest resistance is $0.01>0$, so the logarithm is well defined and no offset is needed. Under this prior the marginal posterior of each coefficient is a Student-$t$ centred at the OLS estimate $\hat\beta$, and the **posterior credible intervals are numerically identical to the OLS confidence intervals**. We can therefore read the full Bayesian inference straight off `lm()`: point estimates from `coef()`, 95% credible intervals from `confint()`, and any posterior probability (such as $P(\beta_F>4)$) from the Student-$t$. The goal is to verify the physical expectation $\beta_F\approx 4$.

```{r loglog-fit}
set.seed(2026)
yh$log_res <- log(yh$resistance)   # no offset: min(resistance) = 0.01 > 0
yh$log_Fr  <- log(yh$Froude)

fit <- lm(log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam, data = yh)

## reference prior => credible intervals = OLS confidence intervals
round(cbind(estimate = coef(fit), confint(fit)), 4)
cat("R^2 =", round(summary(fit)$r.squared, 4),
    "  residual sd =", round(summary(fit)$sigma, 4), "\n")
```

The posterior of the Froude exponent is sharply concentrated around 4.69, well above the textbook value of 4 -- the empirical power law is slightly steeper than $Fn^4$.

```{r loglog-betaF}
cf <- summary(fit)$coefficients
bF <- cf["log_Fr", "Estimate"]; se <- cf["log_Fr", "Std. Error"]; df <- fit$df.residual
ci <- confint(fit)["log_Fr", ]
## beta_F | y ~ t_df centred at bF with scale se  =>  P(beta_F > 4)
cat(sprintf("beta_F: mean = %.3f, 95%% CI = [%.3f, %.3f], P(beta_F > 4) = %.3f\n",
            bF, ci[1], ci[2], pt((bF - 4) / se, df = df)))
```

None of the five hull form factors is individually credible once `Froude` is in the model (every 95% interval contains zero); this redundancy is addressed by the BAS/BMA model selection in part 2.

```{r loglog-fit-plot, fig.width=7, fig.height=4.6}
## log-log scatter with posterior-mean line and 95% credible band
##   (= OLS confidence band under the reference prior)
hull_means <- colMeans(yh[, c("LCB","prismatic","L_disp","beam_draught","L_beam")])
xg   <- seq(min(yh$log_Fr), max(yh$log_Fr), length.out = 120)
band <- predict(fit, data.frame(log_Fr = xg, as.list(hull_means)), interval = "confidence")
band_col <- rgb(.85, .1, .1, .30)
plot(yh$log_Fr, yh$log_res, pch = 16, cex = 0.6, col = rgb(0,0,0,0.45),
     xlab = "log(Froude)", ylab = "log(resistance)", main = "Power-law fit")
polygon(c(xg, rev(xg)), c(band[, "lwr"], rev(band[, "upr"])), col = band_col, border = NA)
lines(xg, band[, "fit"], col = "firebrick", lwd = 2.4)
legend("topleft", bty = "n",
       legend = c("data", "posterior mean fit", "95% credible band"),
       pch = c(16, NA, 15), lty = c(NA, 1, NA), lwd = c(NA, 2.4, NA),
       pt.cex = c(1, NA, 1.8), col = c(rgb(0,0,0,0.45), "firebrick", band_col))
```

```{r loglog-betaF-plot, fig.width=7, fig.height=4.6}
## posterior of beta_F: Student-t centred at the OLS estimate
xb <- seq(bF - 4 * se, bF + 4 * se, length.out = 200)
plot(xb, dt((xb - bF) / se, df) / se, type = "l", col = "firebrick", lwd = 2.4,
     main = expression("Posterior of "*beta[F]),
     xlab = expression(beta[F]), ylab = "posterior density")
abline(v = ci, col = "firebrick", lty = 2)
abline(v = 4, col = "navy", lwd = 2)
legend("topright", bty = "n", lwd = c(2.4, 1, 2), lty = c(1, 2, 1),
       col = c("firebrick", "firebrick", "navy"),
       legend = c("posterior", "95% CI", expression(beta[F] == 4)))
```

```{r loglog-diagnostics, fig.width=10, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))
plot(fitted(fit), resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45),
     xlab = "fitted log(resistance)", ylab = "residual", main = "Residuals vs fitted")
abline(h = 0, col = "firebrick"); lines(lowess(fitted(fit), resid(fit)), col = "navy", lwd = 2)
qqnorm(resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45), main = "Normal Q-Q")
qqline(resid(fit), col = "firebrick")
par(op)
```

The residuals show a $\cup$-shaped curvature (the local Froude exponent grows with speed) and a variance that shrinks as the fitted value grows (heteroscedasticity). Both defects motivate the heteroscedasticity-corrected nonlinear model of parts 3-4.
