---
title: "Final Project 2026 — Yacht Hydrodynamic Resistance"
output:
  html_document:
    df_print: paged
---

## IMPORTANT:

read carefully the instruction for the final project in the pdf file

**Project2026_INFO.pdf**

## Yacht hydrodynamic resistance

**Source.** UCI Machine Learning Repository, <https://archive.ics.uci.edu/dataset/243/yacht+hydrodynamics>. Original record: Gerritsma, Onnink and Versluis, *Geometry, resistance and stability of the Delft Systematic Yacht Hull Series*, 1981.

**Data.** The dataset describes 22 different hull forms (sailing yacht), each tested at multiple Froude numbers in a towing tank, producing 308 hull-form $\times$ Froude-number rows.

Variables:

`LCB` : longitudinal position of the center of buoyancy (% length).

`prismatic` : prismatic coefficient (dimensionless).

`L_disp` : length-displacement ratio.

`beam_draught` : beam-draught ratio.

`L_beam` : length-beam ratio.

`Froude` : Froude number (dimensionless, $Fn = v/\sqrt{gL}$), in [0.125, 0.45]. The dominant predictor.

`resistance` : residuary resistance per unit weight of displacement (target variable).

**Task.** The physics of the problem says that for sailing yachts the residuary resistance follows a power law of the Froude number with exponent close to 4 in the bulk regime. A good Bayesian project explores this:

1.  **Log-log regression.** Fit $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude}) + \sum_k \beta_k\,\mathrm{hull}_k + \varepsilon\,.$$ Verify that $\hat\beta_F \approx 4$.

2.  **BAS / BMA.** Use `bas.lm` to enumerate the $2^5$ submodels over the five hull-geometry covariates and identify which of them (LCB, prismatic, L/disp, B/draught, L/beam) carry signal once `Froude` is in the model.

3.  **Correct for heteroscedasticity.** Train the nonlinear model $$\mathrm{resistance} = k\,\mathrm{Froude}^{\gamma_F}\,\exp\left\{\sum_k \beta_k\,\mathrm{hull}_k \right\} + \varepsilon$$ where only the form factors that appear relevant in the model selection phase are kept for the nonlinear modelling. Compare the results with the log-log regression and discuss the impact of the heteroscedasticity correction on the inference on $\gamma_F$ and on the hull-geometry covariates.

4.  Plot the posterior predictive curves and compare the curves with or without the treatment of the heteroscedasticity and compare the curves with the data.

```{r}
rm(list=ls())
# Yacht hydrodynamics — UCI #243
yh = read.csv("data/yacht_hydro.csv")
dim(yh)
head(yh)
summary(yh)

# The power-law in Fn is best seen on log-log axes;
# resistance > 0 everywhere (min = 0.01), so plain log() needs no offset
plot(log(yh$Froude), log(yh$resistance),
     pch = 16, cex = 0.5,
     xlab = "log(Froude)", ylab = "log(resistance)",
     main = "Yacht hydrodynamic resistance — power-law scaling")
abline(lm(log(resistance) ~ log(Froude), data = yh),
       col = "red", lwd = 2)
```

## 1. Log-log regression (Bayesian, reference prior)

We fit the log-log model $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude})
  + \sum_{k=1}^{5}\beta_k\,\mathrm{hull}_k + \varepsilon,
  \qquad \varepsilon \sim \mathcal N(0,\sigma^2),$$ with the non-informative reference (Jeffreys) prior $p(\beta,\sigma^2)\propto 1/\sigma^2$. The smallest resistance is $0.01>0$, so the logarithm is well defined and no offset is needed. Under this prior the posterior is Normal-Inverse-Gamma and we sample it by Monte Carlo. The goal is to verify the physical expectation $\beta_F\approx 4$.

```{r loglog-setup}
set.seed(2026)
yh$log_res <- log(yh$resistance)   # no offset: min(resistance) = 0.01 > 0
yh$log_Fr  <- log(yh$Froude)

predictors <- c("log_Fr", "LCB", "prismatic", "L_disp", "beam_draught", "L_beam")
form <- as.formula(paste("log_res ~", paste(predictors, collapse = " + ")))

X <- model.matrix(form, data = yh); y <- yh$log_res
n <- nrow(X); p <- ncol(X)

XtX_inv  <- solve(crossprod(X))
beta_hat <- as.vector(XtX_inv %*% crossprod(X, y))   # OLS = posterior mean
names(beta_hat) <- colnames(X)
resid <- y - X %*% beta_hat
s2    <- as.numeric(crossprod(resid)) / (n - p)
```

```{r loglog-mc}
## Monte Carlo draws from the Normal-Inverse-Gamma posterior:
##   sigma^2 | y      ~ Inv-Gamma( (n-p)/2 , (n-p)s2/2 )
##   beta | sigma^2,y ~ N( beta_hat , sigma^2 (X'X)^{-1} )
M <- 20000
sigma2_draws <- (n - p) * s2 / rchisq(M, df = n - p)
L <- chol(XtX_inv)
beta_draws <- matrix(0, M, p, dimnames = list(NULL, colnames(X)))
for (m in seq_len(M)) {
  beta_draws[m, ] <- beta_hat + sqrt(sigma2_draws[m]) * as.vector(crossprod(L, rnorm(p)))
}

post_ci <- t(apply(beta_draws, 2, quantile, probs = c(0.025, 0.975)))
summ <- data.frame(
  mean  = colMeans(beta_draws),
  sd    = apply(beta_draws, 2, sd),
  q2.5  = post_ci[, 1],
  q97.5 = post_ci[, 2],
  P_gt0 = colMeans(beta_draws > 0)
)
round(summ, 4)
cat("R^2 =", round(1 - sum(resid^2)/sum((y - mean(y))^2), 4),
    "  residual sd =", round(sqrt(s2), 4), "\n")
```

The posterior of the Froude exponent is sharply concentrated around 4.69, well above the textbook value of 4 -- the empirical power law is slightly steeper than $Fn^4$.

```{r loglog-betaF}
bF <- beta_draws[, "log_Fr"]
cat(sprintf("beta_F: mean = %.3f, 95%% CI = [%.3f, %.3f], P(beta_F > 4) = %.3f\n",
            mean(bF), quantile(bF, .025), quantile(bF, .975), mean(bF > 4)))
```

None of the five hull form factors is individually credible once `Froude` is in the model (every 95% interval contains zero); this redundancy is addressed by the BAS/BMA model selection in part 2.

```{r loglog-plots, fig.width=10, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))

## (a) log-log scatter with posterior-mean line and 95% credible band
xg <- seq(min(yh$log_Fr), max(yh$log_Fr), length.out = 120)
hull_means <- colMeans(yh[, c("LCB","prismatic","L_disp","beam_draught","L_beam")])
Xg <- cbind(1, xg, matrix(rep(hull_means, each = length(xg)), nrow = length(xg)))
band <- apply(Xg %*% t(beta_draws), 1, quantile, probs = c(0.025, 0.5, 0.975))
plot(yh$log_Fr, yh$log_res, pch = 16, cex = 0.55, col = rgb(0,0,0,0.45),
     xlab = "log(Froude)", ylab = "log(resistance)", main = "Power-law fit")
polygon(c(xg, rev(xg)), c(band[1,], rev(band[3,])), col = rgb(.85,.1,.1,.18), border = NA)
lines(xg, band[2,], col = "firebrick", lwd = 2.2)

## (b) Monte Carlo posterior of beta_F
plot(density(bF), col = "firebrick", lwd = 2, main = expression("Posterior of "*beta[F]),
     xlab = expression(beta[F]))
abline(v = quantile(bF, c(.025, .975)), col = "firebrick", lty = 2)
abline(v = 4, col = "navy", lwd = 2)
par(op)
```

```{r loglog-diagnostics, fig.width=10, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))
fitted_vals <- as.vector(X %*% beta_hat)
plot(fitted_vals, resid, pch = 16, cex = 0.5, col = rgb(0,0,0,0.45),
     xlab = "fitted log(resistance)", ylab = "residual", main = "Residuals vs fitted")
abline(h = 0, col = "firebrick"); lines(lowess(fitted_vals, resid), col = "navy", lwd = 2)
qqnorm(resid, pch = 16, cex = 0.5, col = rgb(0,0,0,0.45), main = "Normal Q-Q")
qqline(resid, col = "firebrick")
par(op)
```

The residuals show a $\cup$-shaped curvature (the local Froude exponent grows with speed) and a variance that shrinks as the fitted value grows (heteroscedasticity). Both defects motivate the heteroscedasticity-corrected nonlinear model of parts 3-4. All figures above are produced directly by the code chunks in this notebook.

## 1.1 Robustness to an informative prior

The reference prior above is deliberately *objective*: it lets the data alone estimate $\beta_F$, and we then **verify** the result against the physical value 4. A complementary, more subjective-Bayesian view is to **encode** the physics directly as an informative prior $\beta_F\sim\mathcal N(4,\tau^2)$ (vague on the remaining coefficients and on $\sigma^2$) and ask how strongly the prior would have to insist on 4 to override the data. The full conditionals are closed form, so a short Gibbs sampler suffices; we sweep the prior standard deviation $\tau$.

```{r informative-prior}
## beta | sigma^2, y  ~ N( V (S0^{-1} m0 + X'y/sigma^2), V ),  V = (S0^{-1} + X'X/sigma^2)^{-1}
## sigma^2 | beta, y  ~ Inv-Gamma( a0 + n/2 , b0 + ||y - X beta||^2 / 2 )
set.seed(2026)
fr  <- which(colnames(X) == "log_Fr")
XtX <- crossprod(X); Xty <- crossprod(X, y)

gibbs_betaF <- function(tau, n_iter = 8000, burn = 2000) {
  m0  <- rep(0, p); m0[fr] <- 4                  # prior mean: physics only on beta_F
  psd <- rep(1000, p); psd[fr] <- tau            # prior sd: vague except beta_F
  S0_inv <- diag(1 / psd^2); a0 <- 1e-3; b0 <- 1e-3
  beta <- beta_hat; sig2 <- s2; keep <- numeric(n_iter - burn)
  for (it in seq_len(n_iter)) {
    V    <- solve(S0_inv + XtX / sig2)
    mu   <- V %*% (S0_inv %*% m0 + Xty / sig2)
    beta <- as.vector(mu + t(chol(V)) %*% rnorm(p))
    sig2 <- 1 / rgamma(1, a0 + n / 2, b0 + sum((y - X %*% beta)^2) / 2)
    if (it > burn) keep[it - burn] <- beta[fr]
  }
  keep
}

taus  <- c(0.05, 0.1, 0.2, 0.5)
draws <- lapply(taus, gibbs_betaF)
tab <- t(sapply(draws, function(d)
  c(post_mean = mean(d), q2.5 = quantile(d, .025),
    q97.5 = quantile(d, .975), P_gt4 = mean(d > 4))))
rownames(tab) <- sprintf("tau = %.2f", taus)
round(rbind(tab,
            reference = c(mean(bF), quantile(bF, .025),
                          quantile(bF, .975), mean(bF > 4))), 3)
```

```{r informative-prior-plot, fig.width=7, fig.height=4.2}
cols <- c("#d7301f", "#fc8d59", "#f1a340", "#2b8cbe")
plot(density(bF), lwd = 2.6, col = "grey30", xlim = c(3.9, 4.9),
     main = expression("Posterior of "*beta[F]*" vs prior strength"),
     xlab = expression(beta[F]))
for (i in seq_along(taus)) lines(density(draws[[i]]), col = cols[i], lwd = 2)
abline(v = 4, col = "navy", lty = 2, lwd = 2)
legend("topright", bty = "n", lwd = 2, col = c("grey30", cols),
       legend = c("reference  (1/sigma^2)", sprintf("N(4, %.2f^2)", taus)))
```

With a diffuse prior the analysis reproduces the reference result ($\beta_F\approx 4.69$). As the prior centered at 4 is tightened, the posterior mean is pulled toward 4 -- but only a near-dogmatic prior ($\tau=0.05$, a prior 95% interval of roughly $[3.9,4.1]$) drags it as far as $4.33$, and even then $P(\beta_F>4)=1$. For any moderately diffuse physical prior ($\tau\ge 0.1$) the likelihood dominates and the posterior stays in the range $4.56$--$4.69$. The conclusion that the empirical exponent is genuinely steeper than the textbook $Fn^4$ is therefore **robust to the prior**: it is a feature of the data ($n=308$), not an artifact of the non-informative choice. This is also a clean illustration of prior--data conflict -- the likelihood is informative enough to overwhelm all but an unreasonably confident prior.
