---
title: "Final Project 2026 — Yacht Hydrodynamic Resistance"
output:
  html_document:
    df_print: paged
---

## IMPORTANT:

read carefully the instruction for the final project in the pdf file

**Project2026_INFO.pdf**

## Yacht hydrodynamic resistance

**Source.** UCI Machine Learning Repository, <https://archive.ics.uci.edu/dataset/243/yacht+hydrodynamics>. Original record: Gerritsma, Onnink and Versluis, *Geometry, resistance and stability of the Delft Systematic Yacht Hull Series*, 1981.

**Data.** The dataset describes 22 different hull forms (sailing yacht), each tested at multiple Froude numbers in a towing tank, producing 308 hull-form $\times$ Froude-number rows.

Variables:

`LCB` : longitudinal position of the center of buoyancy (% length).

`prismatic` : prismatic coefficient (dimensionless).

`L_disp` : length-displacement ratio.

`beam_draught` : beam-draught ratio.

`L_beam` : length-beam ratio.

`Froude` : Froude number (dimensionless, $Fn = v/\sqrt{gL}$), in [0.125, 0.45]. The dominant predictor.

`resistance` : residuary resistance per unit weight of displacement (target variable).

**Task.** The physics of the problem says that for sailing yachts the residuary resistance follows a power law of the Froude number with exponent close to 4 in the bulk regime. A good Bayesian project explores this:

1.  **Log-log regression.** Fit $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude}) + \sum_k \beta_k\,\mathrm{hull}_k + \varepsilon\,.$$ Verify that $\hat\beta_F \approx 4$.

2.  **BAS / BMA.** Use `bas.lm` to enumerate the $2^5$ submodels over the five hull-geometry covariates and identify which of them (LCB, prismatic, L/disp, B/draught, L/beam) carry signal once `Froude` is in the model.

3.  **Correct for heteroscedasticity.** Train the nonlinear model $$\mathrm{resistance} = k\,\mathrm{Froude}^{\gamma_F}\,\exp\left\{\sum_k \beta_k\,\mathrm{hull}_k \right\} + \varepsilon$$ where only the form factors that appear relevant in the model selection phase are kept for the nonlinear modelling. Compare the results with the log-log regression and discuss the impact of the heteroscedasticity correction on the inference on $\gamma_F$ and on the hull-geometry covariates.

4.  Plot the posterior predictive curves and compare the curves with or without the treatment of the heteroscedasticity and compare the curves with the data.

```{r}
rm(list=ls())
# Yacht hydrodynamics — UCI #243
yh = read.csv("data/yacht_hydro.csv")
dim(yh)
head(yh)
summary(yh)

# The power-law in Fn is best seen on log-log axes;
# resistance > 0 everywhere (min = 0.01), so plain log() needs no offset
plot(log(yh$Froude), log(yh$resistance),
     pch = 16, cex = 0.5,
     xlab = "log(Froude)", ylab = "log(resistance)",
     main = "Yacht hydrodynamic resistance — power-law scaling")
abline(lm(log(resistance) ~ log(Froude), data = yh),
       col = "red", lwd = 2)
```

## 1. Log-log regression (Bayesian, reference prior)

We fit the log-log model $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude})
  + \sum_{k=1}^{5}\beta_k\,\mathrm{hull}_k + \varepsilon,
  \qquad \varepsilon \sim \mathcal N(0,\sigma^2),$$ with the non-informative reference (Jeffreys) prior $\pi(\beta\mid\sigma^2)\propto 1$, $\pi(\sigma^2)\propto 1/\sigma^2$. The smallest resistance is $0.01>0$, so the logarithm is well defined and no offset is needed. Under this prior the marginal posterior of each coefficient is a Student-$t$ centred at the OLS estimate $\hat\beta$, and the **posterior credible intervals are numerically identical to the OLS confidence intervals**. We can therefore read the full Bayesian inference straight off `lm()`: point estimates from `coef()`, 95% credible intervals from `confint()`, and any posterior probability (such as $P(\beta_F>4)$) from the Student-$t$. The goal is to verify the physical expectation $\beta_F\approx 4$.

```{r loglog-fit}
set.seed(2026)
yh$log_res <- log(yh$resistance)   # no offset: min(resistance) = 0.01 > 0
yh$log_Fr  <- log(yh$Froude)

fit <- lm(log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam, data = yh)

## reference prior => credible intervals = OLS confidence intervals
round(cbind(estimate = coef(fit), confint(fit)), 4)
cat("R^2 =", round(summary(fit)$r.squared, 4),
    "  residual sd =", round(summary(fit)$sigma, 4), "\n")
```

The posterior of the Froude exponent is sharply concentrated around 4.69, well above the textbook value of 4 -- the empirical power law is slightly steeper than $Fn^4$.

```{r loglog-betaF}
cf <- summary(fit)$coefficients
bF <- cf["log_Fr", "Estimate"]; se <- cf["log_Fr", "Std. Error"]; df <- fit$df.residual
ci <- confint(fit)["log_Fr", ]
## beta_F | y ~ t_df centred at bF with scale se  =>  P(beta_F > 4)
cat(sprintf("beta_F: mean = %.3f, 95%% CI = [%.3f, %.3f], P(beta_F > 4) = %.3f\n",
            bF, ci[1], ci[2], pt((bF - 4) / se, df = df)))
```

None of the five hull form factors is individually credible once `Froude` is in the model (every 95% interval contains zero); this redundancy is addressed by the BAS/BMA model selection in part 2.

```{r loglog-fit-plot, fig.width=7, fig.height=4.6}
## log-log scatter with posterior-mean line and 95% credible band
##   (= OLS confidence band under the reference prior)
hull_means <- colMeans(yh[, c("LCB","prismatic","L_disp","beam_draught","L_beam")])
xg   <- seq(min(yh$log_Fr), max(yh$log_Fr), length.out = 120)
band <- predict(fit, data.frame(log_Fr = xg, as.list(hull_means)), interval = "confidence")
band_col <- rgb(.85, .1, .1, .30)
plot(yh$log_Fr, yh$log_res, pch = 16, cex = 0.6, col = rgb(0,0,0,0.45),
     xlab = "log(Froude)", ylab = "log(resistance)", main = "Power-law fit")
polygon(c(xg, rev(xg)), c(band[, "lwr"], rev(band[, "upr"])), col = band_col, border = NA)
lines(xg, band[, "fit"], col = "firebrick", lwd = 2.4)
legend("topleft", bty = "n",
       legend = c("data", "posterior mean fit", "95% credible band"),
       pch = c(16, NA, 15), lty = c(NA, 1, NA), lwd = c(NA, 2.4, NA),
       pt.cex = c(1, NA, 1.8), col = c(rgb(0,0,0,0.45), "firebrick", band_col))
```

```{r loglog-betaF-plot, fig.width=7, fig.height=4.6}
## posterior of beta_F: Student-t centred at the OLS estimate
xb <- seq(bF - 4 * se, bF + 4 * se, length.out = 200)
plot(xb, dt((xb - bF) / se, df) / se, type = "l", col = "firebrick", lwd = 2.4,
     main = expression("Posterior of "*beta[F]),
     xlab = expression(beta[F]), ylab = "posterior density")
abline(v = ci, col = "firebrick", lty = 2)
abline(v = 4, col = "navy", lwd = 2)
legend("topright", bty = "n", lwd = c(2.4, 1, 2), lty = c(1, 2, 1),
       col = c("firebrick", "firebrick", "navy"),
       legend = c("posterior", "95% CI", expression(beta[F] == 4)))
```

```{r loglog-diagnostics, fig.width=10, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))
plot(fitted(fit), resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45),
     xlab = "fitted log(resistance)", ylab = "residual", main = "Residuals vs fitted")
abline(h = 0, col = "firebrick"); lines(lowess(fitted(fit), resid(fit)), col = "navy", lwd = 2)
qqnorm(resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45), main = "Normal Q-Q")
qqline(resid(fit), col = "firebrick")
par(op)
```

The residuals show a $\cup$-shaped curvature (the local Froude exponent grows with speed) and a variance that shrinks as the fitted value grows (heteroscedasticity). Both defects motivate the heteroscedasticity-corrected nonlinear model of parts 3-4.

### Sensitivity check: independent informative priors (Gibbs sampler)

The reference prior above is deliberately *objective*: the data alone estimate $\beta_F$, and we then **verify** the result against the physical value 4. A complementary check *encodes* the physics in the prior and asks how strongly it must insist on 4 before it overrides the data. We replace the reference prior with **independent proper priors** $$\beta_F\sim\mathcal N(4,\tau^2),\qquad \beta_j\sim\mathcal N(0,100^2)\ \ (j\neq F),\qquad \sigma^2\sim\text{Inv-Gamma}(10^{-3},10^{-3}),$$ all components mutually independent. Because $\beta$ and $\sigma^2$ are now independent a priori (unlike the conjugate Normal-Inverse-Gamma family, where $\mathrm{Var}(\beta\mid\sigma^2)\propto\sigma^2$), the joint posterior has no closed form -- but both full conditionals are standard, so we sample it with a **Gibbs sampler** and sweep the prior strength $\tau\in\{0.05,0.1,0.2,0.5\}$.

```{r informative-prior-gibbs}
## Full conditionals of the independent (semi-conjugate) prior:
##   beta | sigma^2, y    ~ N( V (S0^{-1} m0 + X'y / sigma^2), V ),
##                          with V = (S0^{-1} + X'X / sigma^2)^{-1}
##   sigma^2 | beta, y    ~ Inv-Gamma( a0 + n/2, b0 + ||y - X beta||^2 / 2 )
set.seed(2026)
X <- model.matrix(fit); y <- yh$log_res
n <- nrow(X); p <- ncol(X)
fr  <- which(colnames(X) == "log_Fr")
XtX <- crossprod(X); Xty <- crossprod(X, y)

gibbs_betaF <- function(tau, n_iter = 10000, burn = 2000) {
  m0  <- rep(0, p); m0[fr] <- 4            # physics prior centred at 4 on beta_F only
  psd <- rep(100, p); psd[fr] <- tau       # vague-but-proper N(0, 100^2) elsewhere
  S0_inv <- diag(1 / psd^2)
  a0 <- 1e-3; b0 <- 1e-3
  beta <- coef(fit); sig2 <- summary(fit)$sigma^2
  keep <- numeric(n_iter - burn)
  for (it in seq_len(n_iter)) {
    V    <- chol2inv(chol(S0_inv + XtX / sig2))
    mu   <- V %*% (S0_inv %*% m0 + Xty / sig2)
    beta <- as.vector(mu + t(chol(V)) %*% rnorm(p))
    sig2 <- 1 / rgamma(1, a0 + n / 2, b0 + sum((y - X %*% beta)^2) / 2)
    if (it > burn) keep[it - burn] <- beta[fr]
  }
  keep
}

taus  <- c(0.05, 0.1, 0.2, 0.5)
draws <- lapply(taus, gibbs_betaF)

tab <- t(sapply(draws, function(d)
  c(mean = mean(d), q2.5 = unname(quantile(d, .025)),
    q97.5 = unname(quantile(d, .975)), P_gt4 = mean(d > 4))))
rownames(tab) <- sprintf("tau = %.2f", taus)
## reference row: closed-form Student-t posterior (= OLS confint)
ref <- c(bF, ci[1], ci[2], 1 - pt((4 - bF) / se, df))
round(rbind(tab, reference = ref), 3)

## crude ACF-based effective sample size, as an MCMC sanity check
ess <- sapply(draws, function(d) {
  ac <- acf(d, lag.max = 50, plot = FALSE)$acf[-1]
  length(d) / (1 + 2 * sum(ac[ac > 0]))
})
cat("approx ESS for beta_F:", round(ess), "\n")
```

```{r informative-prior-plot, fig.width=7, fig.height=4.4}
draw_overlay <- function() {
  cols <- c("#d7301f", "#fc8d59", "#f1a340", "#2b8cbe")
  xb <- seq(3.9, 4.95, length.out = 400)
  plot(xb, dt((xb - bF) / se, df) / se, type = "l", lwd = 2.6, col = "grey30",
       main = expression("Posterior of " * beta[F] * " vs prior strength"),
       xlab = expression(beta[F]), ylab = "posterior density",
       ylim = c(0, max(sapply(draws, function(d) max(density(d)$y)),
                       dt(0, df) / se)))
  for (i in seq_along(taus)) lines(density(draws[[i]]), col = cols[i], lwd = 2)
  abline(v = 4, col = "navy", lty = 2, lwd = 2)
  legend("topleft", bty = "n", lwd = 2, col = c("grey30", cols),
         legend = c(expression("reference " * pi(sigma^2) %prop% sigma^-2),
                    sprintf("N(4, %.2f²)", taus)))
}
draw_overlay()
## keep the figure used by the report in sync with the notebook
pdf("figures/posterior_betaF_prior.pdf", width = 7, height = 4.4)
draw_overlay(); invisible(dev.off())
```

With $\tau=0.5$ the informative analysis reproduces the reference result ($\beta_F\approx4.69$). Tightening the prior pulls the posterior mean toward 4, but only the near-dogmatic $\tau=0.05$ (prior 95% interval roughly $[3.9,4.1]$) drags it as far as $4.33$ -- and even then $P(\beta_F>4\mid y)=1$. For any moderately diffuse physical prior ($\tau\ge0.1$) the likelihood dominates and the posterior mean stays in $4.56$--$4.68$. The conclusion that the empirical exponent is genuinely steeper than $Fn^4$ is therefore **robust to the prior**: it is a feature of the $n=308$ observations, not an artifact of the non-informative choice -- a clean illustration of prior--data conflict, where an informative likelihood overwhelms all but an unreasonably confident prior.

## 2. BAS / BMA model selection over hull covariates

The first model already shows the main physical effect: `log_Fr` is overwhelmingly important. The second question is more subtle: **after Froude is already in the model**, do any of the five hull-geometry variables still add useful information?

We answer this with **Bayesian Adaptive Sampling / Bayesian Model Averaging**. In this project the model space is small: there are five optional hull covariates, so there are only $2^5 = 32$ possible submodels. We keep `log_Fr` in every model and let `bas.lm` enumerate every combination of

`LCB`, `prismatic`, `L_disp`, `beam_draught`, and `L_beam`.

The key output is the **posterior inclusion probability** (PIP): the posterior probability that a variable belongs in the regression model. A PIP near 1 means strong evidence that the variable carries signal; a PIP near 0.5 is weak/ambiguous; a small PIP means little evidence once the other variables and Froude are considered.

```{r bas-bma, message=FALSE, warning=FALSE}
if (!requireNamespace("BAS", quietly = TRUE)) {
  stop("Package 'BAS' is required. Install it with install.packages('BAS') before knitting.")
}
library(BAS)

# Optional hull-geometry covariates. These are the five variables requested in the project.
hull_covariates <- c("LCB", "prismatic", "L_disp", "beam_draught", "L_beam")

# Sanity check: fail loudly if a requested covariate is missing or misspelled.
stopifnot(all(c("log_res", "log_Fr", hull_covariates) %in% names(yh)))

# In the model-selection step, log_Fr must be present in every model.
# The five hull covariates are optional, so bas.lm enumerates 2^5 = 32 submodels.
bas_fit <- bas.lm(
  log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam,
  data = yh,
  prior = "BIC",
  modelprior = uniform(),
  method = "deterministic",
  n.models = 2^length(hull_covariates),
  include.always = ~ log_Fr,
  force.heredity = FALSE
)

# Check that the model space has really been fully enumerated.
n_models_expected <- 2^length(hull_covariates)
n_models_observed <- length(bas_fit$postprobs)
cat("Number of models enumerated:", n_models_observed, "\n")
stopifnot(n_models_observed == n_models_expected)

# BAS may store the selected variables either as a list of indices or as a matrix,
# depending on the package version. The following code converts both cases into
# a stable 0/1 model-inclusion matrix.
term_names <- c("(Intercept)", "log_Fr", hull_covariates)

make_which_matrix <- function(which_obj, term_names) {
  if (is.list(which_obj) && !is.matrix(which_obj)) {
    out <- t(sapply(which_obj, function(idx) {
      # BAS indexes terms from 0: 0 = intercept, 1 = log_Fr, 2:6 = hull covariates.
      as.integer((seq_along(term_names) - 1) %in% idx)
    }))
    colnames(out) <- term_names
    return(out)
  }

  out <- as.matrix(which_obj)
  if (ncol(out) == length(term_names)) {
    colnames(out) <- term_names
  } else if (ncol(out) == length(term_names) - 1) {
    colnames(out) <- term_names[-1]
  } else {
    stop("Unexpected format for bas_fit$which. Inspect str(bas_fit$which).")
  }
  out
}

model_matrix <- as.data.frame(make_which_matrix(bas_fit$which, term_names))

# Normalise posterior model probabilities and attach them to the model matrix.
post_prob <- bas_fit$postprobs / sum(bas_fit$postprobs)
model_matrix$post_prob <- post_prob

# Check that log_Fr is indeed included in all models and that the five hull terms exist.
stopifnot(all(model_matrix$log_Fr == 1))
stopifnot(all(hull_covariates %in% names(model_matrix)))

# Posterior inclusion probabilities computed directly from the enumerated models.
# This avoids relying on version-dependent names in bas_fit$probne0.
pip_hull <- sapply(hull_covariates, function(v) {
  sum(model_matrix[[v]] * model_matrix$post_prob)
})

pip_sorted <- sort(pip_hull, decreasing = TRUE)
pip_table <- data.frame(
  covariate = names(pip_sorted),
  PIP = round(as.numeric(pip_sorted), 3),
  evidence = ifelse(pip_sorted >= 0.75, "strong",
                    ifelse(pip_sorted >= 0.50, "moderate", "weak")),
  row.names = NULL
)

pip_table

# Explicit answer to the project question:
# "which of LCB, prismatic, L/disp, B/draught, L/beam carry signal once Froude is in the model?"
label_map <- c(
  LCB = "LCB",
  prismatic = "prismatic",
  L_disp = "L/disp",
  beam_draught = "B/draught",
  L_beam = "L/beam"
)

signal_table <- data.frame(
  requested_name = unname(label_map[pip_table$covariate]),
  R_variable = pip_table$covariate,
  PIP = pip_table$PIP,
  evidence = pip_table$evidence,
  answer = ifelse(
    pip_table$evidence == "moderate",
    "carries moderate signal",
    "weak evidence; not retained as carrying signal"
  ),
  row.names = NULL
)

signal_table

best_covariate <- pip_table$covariate[1]
best_requested_name <- unname(label_map[best_covariate])
best_pip <- pip_table$PIP[1]

cat("Largest PIP:", best_requested_name, "(", best_covariate, ") =", best_pip, "\n")
cat(
  "Answer:",
  best_requested_name, "(", best_covariate, ") is the only hull-geometry covariate with moderate evidence of carrying signal once Froude is included; the remaining hull covariates show weak evidence.\n"
)
```

**Explicit answer to the project question.** The covariate that carries signal once `Froude` is in the model is **B/draught**, implemented in the data as `beam_draught`. In the rendered BAS/BMA output its PIP is about **0.565**, so the evidence is **moderate** rather than strong. The other four hull-geometry covariates -- `LCB`, `prismatic`, `L_disp` (L/disp), and `L_beam` (L/beam) -- have PIP values below 0.50, so they show weak evidence and are not retained as variables that clearly carry signal.

```{r bas-pip-plot, fig.width=7, fig.height=4.6}
barplot(
  rev(pip_sorted),
  horiz = TRUE,
  las = 1,
  xlim = c(0, 1),
  xlab = "Posterior inclusion probability",
  main = "BAS/BMA posterior inclusion probabilities"
)
abline(v = 0.5, lty = 2)
text(
  x = rev(pip_sorted) + 0.04,
  y = seq_along(pip_sorted),
  labels = round(rev(pip_sorted), 3),
  cex = 0.8
)
```

For transparency, we also print the most probable models. The formula below should be read as: `log_Fr` is always included; the listed hull variables are the extra terms selected in that submodel.

```{r bas-top-models}
top_models <- model_matrix[
  order(-model_matrix$post_prob),
  c(hull_covariates, "post_prob")
]

top_models$model <- apply(
  top_models[, hull_covariates, drop = FALSE],
  1,
  function(z) {
    chosen <- hull_covariates[as.logical(z)]
    if (length(chosen) == 0) {
      "log_Fr only"
    } else {
      paste("log_Fr +", paste(chosen, collapse = " + "))
    }
  }
)

top_models <- top_models[, c("model", "post_prob")]
head(transform(top_models, post_prob = round(post_prob, 3)), 10)
```

Finally, we report the Bayesian model-averaged coefficient summary. These coefficients are averaged over all 32 models, so variables with low inclusion probability are automatically shrunk toward zero in the BMA estimate.

```{r bas-coef-summary}
bma_coef <- coef(bas_fit)
print(bma_coef)
```

### Prior sensitivity analysis for BAS/BMA

The project instructions ask for posterior analysis and, where appropriate, sensitivity analysis. Since the baseline PIP values are not high, we should check whether the conclusion about hull-geometry covariates depends strongly on the prior used in `bas.lm`. Therefore, we repeat the same BAS/BMA enumeration with three standard choices from the linear-regression/model-selection material:

-   `prior = "BIC"`: the baseline reference/BIC approximation used above;
-   `prior = "g-prior"`, with `alpha = n`: Zellner's $g$-prior with $g=n$, a common weakly informative choice;
-   `prior = "JZS"`: Jeffreys-Zellner-Siow prior, a heavy-tailed alternative to a fixed $g$-prior.

In all cases we keep the same model prior, `modelprior = uniform()`, and we still force `log_Fr` to be present in every model. Therefore, each sensitivity run answers the same question: **among `LCB`, `prismatic`, `L_disp`, `beam_draught`, and `L_beam`, which variables carry signal once Froude is already included?**

```{r bas-prior-sensitivity, message=FALSE, warning=FALSE}
fit_bas_prior <- function(prior_name, alpha = NULL) {
  args <- list(
    formula = log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam,
    data = yh,
    prior = prior_name,
    modelprior = uniform(),
    method = "deterministic",
    n.models = n_models_expected,
    include.always = ~ log_Fr,
    force.heredity = FALSE
  )
  if (!is.null(alpha)) {
    args$alpha <- alpha
  }
  do.call(bas.lm, args)
}

extract_pip_table <- function(fit_obj, prior_label) {
  mm <- as.data.frame(make_which_matrix(fit_obj$which, term_names))
  pp <- fit_obj$postprobs / sum(fit_obj$postprobs)
  mm$post_prob <- pp

  # Same checks as in the baseline analysis.
  stopifnot(length(pp) == nrow(mm))
  stopifnot(length(pp) == n_models_expected)
  stopifnot(all(mm$log_Fr == 1))

  pip <- sapply(hull_covariates, function(v) {
    sum(mm[[v]] * mm$post_prob)
  })

  out <- data.frame(
    prior = prior_label,
    requested_name = unname(label_map[names(pip)]),
    R_variable = names(pip),
    PIP = round(as.numeric(pip), 3),
    evidence = ifelse(pip >= 0.75, "strong",
                      ifelse(pip >= 0.50, "moderate", "weak")),
    row.names = NULL
  )
  out[order(-out$PIP), ]
}

bas_fits_sensitivity <- list(
  "BIC / reference" = bas_fit,
  "g-prior (g = n)" = fit_bas_prior("g-prior", alpha = nrow(yh)),
  "JZS" = fit_bas_prior("JZS")
)

sensitivity_table <- do.call(
  rbind,
  lapply(names(bas_fits_sensitivity), function(pr) {
    extract_pip_table(bas_fits_sensitivity[[pr]], pr)
  })
)
rownames(sensitivity_table) <- NULL

# Full sensitivity table: PIP for every hull covariate under every prior.
sensitivity_table

# Most supported hull covariate under each prior.
top_by_prior <- do.call(
  rbind,
  lapply(split(sensitivity_table, sensitivity_table$prior), function(d) {
    d[which.max(d$PIP), ]
  })
)
rownames(top_by_prior) <- NULL

top_by_prior

# Variables that reach at least moderate evidence under every prior.
moderate_sets <- lapply(split(sensitivity_table, sensitivity_table$prior), function(d) {
  d$R_variable[d$PIP >= 0.50]
})
robust_moderate <- Reduce(intersect, moderate_sets)
robust_moderate_names <- unname(label_map[robust_moderate])

if (length(robust_moderate) == 0) {
  sensitivity_answer <- paste(
    "No hull covariate reaches PIP >= 0.50 under all priors.",
    "The safest conclusion is that hull effects are weak and prior-sensitive once Froude is included."
  )
} else {
  sensitivity_answer <- paste(
    paste(robust_moderate_names, collapse = ", "),
    "is/are the hull covariate(s) with at least moderate evidence under all priors.",
    "The remaining hull covariates have weak or prior-sensitive support."
  )
}

cat("Sensitivity answer:", sensitivity_answer, "\n")
```

The sensitivity analysis is important because the baseline PIP values are modest. It avoids overclaiming from a single prior. If the same hull covariate remains the top candidate or the only variable with at least moderate inclusion probability across `BIC`, Zellner's `g-prior`, and `JZS`, then it is reasonable to carry that variable forward into the nonlinear model. If the conclusion changes across priors, then the correct interpretation is more cautious: the data do not strongly identify a stable hull-geometry effect once Froude is included.

**Sensitivity-based interpretation.** `r sensitivity_answer`

**Conclusion of part 2.** The BAS/BMA analysis confirms that `log_Fr` is the dominant explanatory variable and that the hull-geometry covariates provide, at most, secondary information once Froude is already included in the model. Under the baseline BIC/reference specification, **B/draught** (`beam_draught`) has the largest posterior inclusion probability, approximately **0.565**, so it is the main candidate among the five hull covariates. However, this is only moderate evidence, not strong evidence.

The prior sensitivity analysis gives a more cautious interpretation. When the BAS/BMA enumeration is repeated using alternative priors, such as Zellner’s `g-prior` and the `JZS` prior, no hull covariate reaches PIP ≥ 0.50 under all prior choices. Therefore, the contribution of the hull-geometry variables is not fully robust to the prior specification.

Overall, the safest conclusion is that **Froude explains the main resistance scaling**, while the hull-geometry effects are weak to moderate and prior-sensitive. Among them, **B/draught** (`beam_draught`) is the most defensible reduced candidate to carry forward into the nonlinear heteroscedasticity-corrected model of part 3, but it should also be compared with a no-hull model as a sensitivity check.

## 3. Heteroscedasticity-corrected nonlinear model (random-walk Metropolis)

The log-log model of part 1 assumes *multiplicative* lognormal errors: back on the original scale it implies a noise standard deviation proportional to the resistance itself, which the residual diagnostics contradicted. Part 3 keeps the same power-law mean but moves the noise to the original scale, **additive and homoscedastic**:

$$\mathrm{resistance}_i \;=\; \mu_i + \varepsilon_i, \qquad \varepsilon_i \sim \mathcal N(0, \sigma^2)
\qquad\Longleftrightarrow\qquad \mathrm{resistance}_i \;\sim\; \mathcal N(\mu_i,\ \sigma^2),$$

with mean $\mu_i = k\,\mathrm{Froude}_i^{\gamma_F}\,\exp\{\beta_B\,\mathrm{bd}_i\}$. For the sampler we rewrite this product of powers as the exponential of a linear expression, using only the identities $a = e^{\log a}$ and $a^b = e^{b \log a}$ (valid because $k > 0$ and $\mathrm{Froude}_i > 0$):

$$\mu_i
= k \cdot \mathrm{Froude}_i^{\gamma_F} \cdot e^{\beta_B\,\mathrm{bd}_i}
= e^{\log k} \cdot e^{\gamma_F \log \mathrm{Froude}_i} \cdot e^{\beta_B\,\mathrm{bd}_i}
= \exp\big\{\log k + \gamma_F \log \mathrm{Froude}_i + \beta_B\,\mathrm{bd}_i\big\}.$$

Next we centre the covariates: writing $\log \mathrm{Froude}_i = m_F + \tilde x_{F,i}$ and $\mathrm{bd}_i = m_B + \tilde x_{B,i}$ (centring constant plus centred variable) and collecting all constants into a single intercept,

$$\mu_i
= \exp\big\{\underbrace{\log k + \gamma_F m_F + \beta_B m_B}_{\textstyle b_0}
  \;+\; \gamma_F\,\tilde x_{F,i} \;+\; \beta_B\,\tilde x_{B,i}\big\}
= \exp\big\{b_0 + \gamma_F\,\tilde x_{F,i} + \beta_B\,\tilde x_{B,i}\big\}.$$

The slopes $\gamma_F$ and $\beta_B$ are untouched by this rewriting; only the intercept is relabelled. After sampling we recover the physical constant by inverting the definition of $b_0$:

$$k \;=\; \exp\big\{b_0 - \gamma_F m_F - \beta_B m_B\big\} \quad \text{(applied to every posterior draw).}$$

Following the BAS result of part 2 we keep only `beam_draught` (bd) and fit two models: **M0** (Froude only) and **M1** (Froude + beam_draught). Priors: $b_0\sim\mathcal N(0,10^2)$, $\gamma_F\sim\mathcal N(4,2^2)$ (weakly informative, centred at the physical value), $\beta_B\sim\mathcal N(0,2^2)$, $1/\sigma^2\sim\mathrm{Gamma}(0.01,0.01)$. As a sensitivity check the $\gamma_F$ prior is also replaced by a vague $\mathcal N(0,10^2)$ and an informative $\mathcal N(4,0.5^2)$.

The parameters enter the mean through an exponential, so the likelihood is not conjugate with any standard prior and the posterior has no closed form. We sample it with a **random-walk Metropolis-within-Gibbs** sampler. At each iteration we propose the whole regression vector $\theta=(b_0,\gamma_F,\beta_B)$ by $$\theta^\star=\theta^{(t)}+\eta,\qquad \eta\sim\mathcal N(0,\mathrm{diag}(s_1^2,\ldots,s_p^2)),$$ and accept/reject it with the usual Metropolis-Hastings posterior ratio. Conditional on $\theta$, the precision $\tau=1/\sigma^2$ is conjugate, so it is updated exactly from its Gamma full conditional. Thus the sampler is random-walk Metropolis for the nonlinear mean parameters, plus a Gibbs update for the precision.

One computational detail concerns the choice of the centring constants $m_F, m_B$. With *ordinary* means, the chains for $b_0$ and $\gamma_F$ mix very poorly: under additive errors the likelihood is dominated by the large-resistance (high-Froude) rows, so changing the slope $\gamma_F$ can be almost perfectly compensated by changing the intercept, and the posterior of $(b_0,\gamma_F)$ becomes a long thin ridge that a random-walk proposal crosses very slowly. We therefore centre at a *weighted* mean with weights proportional to $y_i^2$ which makes $b_0$ and $\gamma_F$ nearly uncorrelated and lets the sampler mix freely. This is purely a reparametrization for the benefit of the sampler (the slopes are untouched and the intercept is back-transformed to $k$ afterwards, so the reported inference is unaffected); we also showed the comparison of the mixing time below.

```{r task3-prep}
y_res <- yh$resistance
n_obs <- length(y_res)

## centre covariates at a y^2-weighted mean: removes the strong posterior
## correlation between b0 and gamma_F that ruins mixing with plain centring
w_info <- y_res^2 / sum(y_res^2)
mF <- sum(w_info * log(yh$Froude))
mB <- sum(w_info * yh$beam_draught)
xF_c <- log(yh$Froude)  - mF
xB_c <- yh$beam_draught - mB

X0 <- cbind(intercept = 1, log_Fr = xF_c)                    # M0: Froude only
X1 <- cbind(X0, beam_draught = xB_c)                         # M1: + beam_draught
```

```{r task3-rwmh-model, message=FALSE, warning=FALSE}
if (!requireNamespace("coda", quietly = TRUE)) {
  stop("Package 'coda' is required. Install it with install.packages('coda') before knitting.")
}
library(coda)

## Log posterior kernel for theta | tau, y.
## The normalizing constants do not matter for the MH acceptance ratio.
logpost_theta <- function(theta, tau, X, prior_mean, prior_sd) {
  eta <- as.vector(X %*% theta)
  if (any(!is.finite(eta)) || any(eta > 50)) {
    return(-Inf)
  }
  mu <- exp(eta)
  if (any(!is.finite(mu))) {
    return(-Inf)
  }
  sum(dnorm(y_res, mu, sd = 1 / sqrt(tau), log = TRUE)) +
    sum(dnorm(theta, prior_mean, prior_sd, log = TRUE))
}

## Random-walk Metropolis-within-Gibbs sampler.
## theta is updated as one block by random-walk MH; tau is updated exactly.
run_rwmh <- function(X, prior_mean, prior_sd, prop_sd, n_chains = 4,
                     n_iter = 100000, burn = 20000, thin = 5,
                     gamma_inits = c(3, 6, 5, 4), seed0 = 2026) {
  p <- ncol(X)
  stopifnot(length(prior_mean) == p, length(prior_sd) == p, length(prop_sd) == p)
  stopifnot(length(gamma_inits) >= n_chains)

  a0 <- 0.01
  b0 <- 0.01
  theta_start <- coef(lm(log(y_res) ~ X - 1))   # crude start from the log fit

  chains <- vector("list", n_chains)
  accept_rate <- numeric(n_chains)
  n_keep <- floor((n_iter - burn) / thin)

  for (ch in seq_len(n_chains)) {
    set.seed(seed0 + ch)
    theta <- theta_start
    theta[2] <- gamma_inits[ch]
    tau <- 1
    current_lp <- logpost_theta(theta, tau, X, prior_mean, prior_sd)
    keep <- matrix(NA_real_, n_keep, p + 2)
    colnames(keep) <- c(paste0("theta[", seq_len(p), "]"), "sigma", "sigma2")

    accepted <- 0L
    k <- 0L
    for (it in seq_len(n_iter)) {
      theta_prop <- theta + rnorm(p, mean = 0, sd = prop_sd)
      prop_lp <- logpost_theta(theta_prop, tau, X, prior_mean, prior_sd)

      if (log(runif(1)) < prop_lp - current_lp) {
        theta <- theta_prop
        current_lp <- prop_lp
        accepted <- accepted + 1L
      }

      resid <- y_res - exp(as.vector(X %*% theta))
      tau <- rgamma(1, shape = a0 + n_obs / 2,
                    rate = b0 + sum(resid^2) / 2)
      current_lp <- logpost_theta(theta, tau, X, prior_mean, prior_sd)

      if (it > burn && (it - burn) %% thin == 0) {
        k <- k + 1L
        sigma2 <- 1 / tau
        keep[k, ] <- c(theta, sqrt(sigma2), sigma2)
      }
    }

    chains[[ch]] <- mcmc(keep)
    accept_rate[ch] <- accepted / n_iter
  }

  out <- mcmc.list(chains)
  attr(out, "accept_rate") <- accept_rate
  attr(out, "proposal_sd") <- prop_sd
  out
}

accept_table <- function(...) {
  fits <- list(...)
  data.frame(
    run = names(fits),
    acceptance_rate = vapply(fits, function(z) {
      paste(round(attr(z, "accept_rate"), 3), collapse = ", ")
    }, character(1)),
    proposal_sd = vapply(fits, function(z) {
      paste(attr(z, "proposal_sd"), collapse = ", ")
    }, character(1)),
    row.names = NULL
  )
}
```

```{r task3-fit, cache=TRUE}
## main prior, 4 chains each
rw_M0 <- run_rwmh(
  X0, prior_mean = c(0, 4), prior_sd = c(10, 2),
  prop_sd = c(0.008, 0.12)
)
rw_M1 <- run_rwmh(
  X1, prior_mean = c(0, 4, 0), prior_sd = c(10, 2, 2),
  prop_sd = c(0.006, 0.095, 0.011)
)

## gamma_F prior sensitivity (M1, 2 chains each)
rw_M1_vague <- run_rwmh(
  X1, prior_mean = c(0, 0, 0), prior_sd = c(10, 10, 2),
  prop_sd = c(0.006, 0.095, 0.011), n_chains = 2
)
rw_M1_inform <- run_rwmh(
  X1, prior_mean = c(0, 4, 0), prior_sd = c(10, 0.5, 2),
  prop_sd = c(0.006, 0.095, 0.011), n_chains = 2
)

accept_table(
  "M0 (main prior)" = rw_M0,
  "M1 (main prior)" = rw_M1,
  "M1 (vague gamma_F prior)" = rw_M1_vague,
  "M1 (informative gamma_F prior)" = rw_M1_inform
)
```

### Convergence diagnostics

MCMC draws are only valid posterior samples after the chains have converged to the target distribution, so before any inference we check five things. (1) **Acceptance rates**: for a random-walk Metropolis update they should be neither close to 0 nor close to 1; values around 0.3--0.5 are acceptable here. (2) **Traceplots**: each parameter's chains should look like stationary noise around a fixed level, with the independently started chains overlapping. (3) **Gelman--Rubin** $\hat R$: the chains were started from deliberately different values of $\gamma_F$ (3, 6, 5, 4); $\hat R$ compares the variance *between* chains to the variance *within* chains, and is $\approx 1$ only when all chains explore the same distribution -- we require $\hat R < 1.01$. (4) **Effective sample size (ESS)**: consecutive MCMC draws are autocorrelated, so the kept MCMC draws carry less information than independent samples; ESS is the equivalent number of independent draws, and we want it in the thousands so that posterior quantiles are stable. (5) **Autocorrelation plots**: the autocorrelation of the chain should decay to zero within a few lags; slow decay explains a low ESS.

```{r task3-diagnostics, fig.width=10, fig.height=6}
par_labels <- c("theta[1]" = "b0", "theta[2]" = "gamma_F",
                "theta[3]" = "beta_B", "sigma" = "sigma")

diag_table <- function(fit, run) {
  keep <- intersect(varnames(fit), names(par_labels))
  rhat <- gelman.diag(fit[, keep], multivariate = FALSE)$psrf[, 1]
  ess  <- effectiveSize(fit[, keep])
  data.frame(run = run, parameter = unname(par_labels[keep]),
             Rhat = round(rhat, 4), ESS = round(ess), row.names = NULL)
}

rbind(diag_table(rw_M0, "M0 (main prior)"),
      diag_table(rw_M1, "M1 (main prior)"),
      diag_table(rw_M1_vague,  "M1 (vague gamma_F prior)"),
      diag_table(rw_M1_inform, "M1 (informative gamma_F prior)"))

## traceplots + autocorrelation for the headline run (M1, main prior)
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
traceplot(rw_M1[, c("theta[1]", "theta[2]", "theta[3]", "sigma")])
par(op)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
acf(as.matrix(rw_M1[, "theta[2]"]), main = "ACF gamma_F (pooled chains)")
acf(as.matrix(rw_M1[, "sigma"]),    main = "ACF sigma (pooled chains)")
par(op)
```

The acceptance rates are in a sensible range, all $\hat R$ are $\approx 1$, and the effective sample sizes are comfortably large, so the chains can be pooled for inference.

### Justifying the weighted centring: the same model with ordinary centring

To show that the weighted centring is doing real work (and nothing else), we re-run the identical model, same priors, same four chains of the same length changing only the centring constants to the ordinary means. Everything that differs between the two runs is therefore attributable to the parametrization alone.

```{r task3-centring-experiment, cache=TRUE}
xF_plain <- log(yh$Froude)  - mean(log(yh$Froude))
xB_plain <- yh$beam_draught - mean(yh$beam_draught)
X1_plain <- cbind(intercept = 1, log_Fr = xF_plain, beam_draught = xB_plain)

rw_M1_plain <- run_rwmh(
  X1_plain, prior_mean = c(0, 4, 0), prior_sd = c(10, 2, 2),
  prop_sd = c(0.006, 0.095, 0.011)
)
```

```{r task3-centring-compare, fig.width=10, fig.height=7}
keep3 <- c("theta[1]", "theta[2]", "theta[3]")
n_tot <- niter(rw_M1) * nchain(rw_M1)     # total kept draws (same in both runs)

mix_stats <- function(fit) {
  rhat <- gelman.diag(fit[, keep3], multivariate = FALSE)$psrf[, 1]
  ess  <- effectiveSize(fit[, keep3])
  cbind(Rhat = round(rhat, 3), ESS = round(ess),
        `draws needed for ESS 1000` = round(1000 * n_tot / ess))
}

cat("ordinary centring (", n_tot, "kept draws ):\n")
mix_stats(rw_M1_plain)
cat("\nweighted centring (same", n_tot, "kept draws ):\n")
mix_stats(rw_M1)

## the cause: posterior correlation between intercept and slope
cor_plain <- cor(as.matrix(rw_M1_plain)[, "theta[1]"],
                 as.matrix(rw_M1_plain)[, "theta[2]"])
cor_wt    <- cor(as.matrix(rw_M1)[, "theta[1]"],
                 as.matrix(rw_M1)[, "theta[2]"])
cat(sprintf("\ncor(b0, gamma_F): ordinary centring %.4f  vs  weighted centring %.4f\n",
            cor_plain, cor_wt))

## same posterior, different Monte Carlo efficiency
round(rbind(`gamma_F, ordinary centring` = summary(rw_M1_plain)$statistics["theta[2]", 1:2],
            `gamma_F, weighted centring` = summary(rw_M1)$statistics["theta[2]", 1:2]), 4)

## visual: gamma_F trace and autocorrelation under the two parametrizations
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
traceplot(rw_M1_plain[, "theta[2]"], main = "gamma_F trace - ordinary centring")
traceplot(rw_M1[, "theta[2]"],       main = "gamma_F trace - weighted centring")
acf(as.matrix(rw_M1_plain[[1]][, "theta[2]"]), lag.max = 200,
    main = "ACF gamma_F - ordinary centring (chain 1)")
acf(as.matrix(rw_M1[[1]][, "theta[2]"]), lag.max = 200,
    main = "ACF gamma_F - weighted centring (chain 1)")
par(op)
```

The comparison makes the case by itself: with ordinary centring the intercept and the Froude exponent are almost perfectly negatively correlated in the posterior, while the weighted centring makes that posterior ridge much closer to axis-aligned. The two parametrizations describe the same mean curve, so they should agree on $\gamma_F$ after convergence; the printed summaries check exactly that. The difference is computational efficiency: the ordinary-centred random walk has stronger autocorrelation and a lower effective sample size, whereas the weighted version reaches the same posterior with much better mixing.

### Posterior analysis

```{r task3-posterior}
## back-transform the centred intercept to the physical k for every draw
post_summary <- function(fit, hull = TRUE) {
  d <- as.matrix(fit)
  k <- exp(d[, "theta[1]"] - d[, "theta[2]"] * mF -
             (if (hull) d[, "theta[3]"] * mB else 0))
  out <- cbind(k = k, gamma_F = d[, "theta[2]"],
               beta_B = if (hull) d[, "theta[3]"] else NULL,
               sigma = d[, "sigma"])
  t(apply(out, 2, function(z)
    c(mean = mean(z), sd = sd(z),
      `2.5%` = unname(quantile(z, .025)), `50%` = unname(quantile(z, .5)),
      `97.5%` = unname(quantile(z, .975)))))
}

cat("M0 (Froude only):\n")
round(post_summary(rw_M0, hull = FALSE), 4)
cat("\nM1 (Froude + beam_draught):\n")
round(post_summary(rw_M1), 4)

gF_draws <- as.matrix(rw_M1)[, "theta[2]"]
bB_draws <- as.matrix(rw_M1)[, "theta[3]"]
cat(sprintf("\nP(gamma_F > 4 | y) = %.4f   P(beta_B < 0 | y) = %.4f\n",
            mean(gF_draws > 4), mean(bB_draws < 0)))
cat("log-log (part 1) for comparison: beta_F =", round(bF, 3),
    ", 95% CI [", round(ci[1], 3), ",", round(ci[2], 3), "]\n")
cat("log-log beam_draught 95% CI: [",
    paste(round(confint(fit)["beam_draught", ], 4), collapse = ", "), "]\n")
```

```{r task3-gamma-plot, fig.width=10, fig.height=4.6}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))

## gamma_F (additive errors) vs beta_F (log-log, closed-form Student-t)
dg <- density(gF_draws)
plot(dg, col = "firebrick", lwd = 2.4, xlim = c(3.9, max(dg$x)),
     main = "Froude exponent under the two error models",
     xlab = expression(gamma[F]), ylab = "posterior density")
curve(dt((x - bF) / se, df) / se, from = 4.3, to = 5.1,
      n = 400, col = "navy", lwd = 2.4, add = TRUE)
abline(v = 4, lty = 2)
legend("top", bty = "n", lwd = c(2.4, 2.4, 1), col = c("firebrick", "navy", "black"),
       lty = c(1, 1, 2),
       legend = c(expression(gamma[F]*" (additive errors, M1)"),
                  expression(beta[F]*" (log-log, part 1)"),
                  expression(gamma[F] == 4)))

## beam_draught effect: credible once the error model is corrected
plot(density(bB_draws), col = "firebrick", lwd = 2.4,
     main = expression("Posterior of "*beta[B]*" (beam-draught)"),
     xlab = expression(beta[B]), ylab = "posterior density")
abline(v = 0, lty = 2)
abline(v = quantile(bB_draws, c(.025, .975)), col = "firebrick", lty = 3)
par(op)
```

```{r task3-sensitivity}
## gamma_F posterior under the three priors (M1)
sens <- rbind(
  "N(4, 2^2)   (main)"        = post_summary(rw_M1)["gamma_F", ],
  "N(0, 10^2)  (vague)"       = post_summary(rw_M1_vague)["gamma_F", ],
  "N(4, 0.5^2) (informative)" = post_summary(rw_M1_inform)["gamma_F", ])
round(sens, 4)
```

**Comparison with the log-log regression.** The heteroscedasticity correction changes the inference dramatically: the Froude exponent jumps from $\beta_F \approx 4.69$ (log-log) to $\gamma_F \approx 7.2$, and the two posteriors do not even overlap. Conversely, the hull covariate becomes more credible, not less: `beam_draught` now has a clearly negative coefficient with a 95% credible interval far from zero, whereas in the log-log model it was indistinguishable from zero. The $\gamma_F$ inference is insensitive to the prior (the vague and main priors agree to two decimals; even the informative $\mathcal N(4,0.5^2)$ prior moves the posterior mean only marginally), so the shift relative to part 1 is a property of the error model, not of the prior.

### Model comparison: does beam_draught earn its place?

WAIC and PSIS-LOO are computed from the MCMC draws. Both use the pointwise log-likelihood matrix $\log p(y_i\mid\theta^{(s)})$ and compare expected out-of-sample predictive accuracy; lower WAIC/LOOIC is better. PSIS-LOO is especially useful because it also reports Pareto-$k$ diagnostics: large $k$ values warn that one or more observations are highly influential and that the leave-one-out approximation may be unreliable.

```{r task3-waic-loo}
if (!requireNamespace("loo", quietly = TRUE)) {
  stop("Package 'loo' is required for PSIS-LOO. Install it with install.packages('loo') before knitting.")
}
library(loo)

log_lik_from_fit <- function(fit, X, n_draws = 4000, seed = 2026) {
  d <- as.matrix(fit)
  set.seed(seed)
  d <- d[sample(nrow(d), n_draws), ]
  th <- d[, paste0("theta[", seq_len(ncol(X)), "]"), drop = FALSE]
  s2 <- d[, "sigma2"]
  mu <- exp(th %*% t(X))                                    # n_draws x n
  dnorm(matrix(y_res, n_draws, n_obs, byrow = TRUE),
        mu, sqrt(s2), log = TRUE)                           # sd recycles by draw
}

ic_from_log_lik <- function(ll, loo_obj = loo(ll, r_eff = NA)) {
  ## WAIC: lppd via log-sum-exp, p_waic = sum of pointwise posterior variances
  mx     <- apply(ll, 2, max)
  lppd   <- sum(mx + log(colMeans(exp(sweep(ll, 2, mx)))))
  p_waic <- sum(apply(ll, 2, var))

  ## PSIS-LOO: approximate leave-one-out CV from the same log-likelihood matrix
  k <- pareto_k_values(loo_obj)

  c(WAIC = -2 * (lppd - p_waic), p_WAIC = p_waic,
    LOOIC = loo_obj$estimates["looic", "Estimate"],
    elpd_LOO = loo_obj$estimates["elpd_loo", "Estimate"],
    p_LOO = loo_obj$estimates["p_loo", "Estimate"],
    max_pareto_k = max(k),
    n_pareto_k_gt_0.7 = sum(k > 0.7))
}

ll_M0 <- log_lik_from_fit(rw_M0, X0)
ll_M1 <- log_lik_from_fit(rw_M1, X1)
loo_M0 <- loo(ll_M0, r_eff = NA)
loo_M1 <- loo(ll_M1, r_eff = NA)

ic <- rbind("M0: Froude only"           = ic_from_log_lik(ll_M0, loo_M0),
            "M1: Froude + beam_draught" = ic_from_log_lik(ll_M1, loo_M1))
round(cbind(ic,
            dWAIC = ic[, "WAIC"] - min(ic[, "WAIC"]),
            dLOOIC = ic[, "LOOIC"] - min(ic[, "LOOIC"])), 1)

loo_compare(list(
  "M0: Froude only" = loo_M0,
  "M1: Froude + beam_draught" = loo_M1
))
```

### Did the heteroscedasticity correction work? Residual diagnostics on the original scale

Part 1 diagnosed the log-log model on the *log* scale; the fair test of the part-3 model is the same diagnostic on the **original** (resistance) scale, where its errors are assumed additive and homoscedastic, $\varepsilon_i\sim\mathcal N(0,\sigma^2)$ with a single constant $\sigma$. Using the posterior-mean fit $\hat\mu_i=\exp(\hat\theta^\top X_i)$ we form the raw residuals $y_i-\hat\mu_i$ and, because the model variance is constant, the standardized residuals $(y_i-\hat\mu_i)/\hat\sigma$. Under a correctly specified homoscedastic model the residual spread should be flat across fitted values (the scale-location curve horizontal) and roughly 95% of the standardized residuals should fall inside a $\pm 2$ band with no funnel against `Froude`.

```{r task3-resid-diag, fig.width=10, fig.height=7.5}
## posterior-mean fit on the original scale (theta, sigma are the constant-variance params)
d_M1      <- as.matrix(rw_M1)
th_hat    <- colMeans(d_M1[, paste0("theta[", seq_len(ncol(X1)), "]")])
sigma_hat <- mean(d_M1[, "sigma"])

fitted_M1 <- as.vector(exp(X1 %*% th_hat))   # fitted resistance, original scale
resid_M1  <- y_res - fitted_M1               # raw residuals
std_M1    <- resid_M1 / sigma_hat            # standardized (model sd is constant)

## quantify heteroscedasticity: residual SD within fitted-value tertiles
ter <- cut(fitted_M1, quantile(fitted_M1, c(0, 1/3, 2/3, 1)),
           include.lowest = TRUE, labels = c("low", "mid", "high"))
sd_by_tertile <- tapply(resid_M1, ter, sd)
cat("residual SD by fitted-value tertile:\n"); print(round(sd_by_tertile, 3))
cat(sprintf("high/low spread ratio = %.1f   (= 1 would be homoscedastic)\n",
            sd_by_tertile["high"] / sd_by_tertile["low"]))
cat(sprintf("cor(|standardized residual|, fitted) = %.3f\n",
            cor(abs(std_M1), fitted_M1)))

op <- par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.2, 1))

## (1) residuals vs fitted
plot(fitted_M1, resid_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
     xlab = "fitted resistance", ylab = "residual",
     main = "Residuals vs fitted (original scale)")
abline(h = 0, col = "firebrick"); lines(lowess(fitted_M1, resid_M1), col = "navy", lwd = 2)

## (2) scale-location: sqrt|standardized residual| vs fitted -> homoscedasticity check
plot(fitted_M1, sqrt(abs(std_M1)), pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
     xlab = "fitted resistance", ylab = "sqrt(|standardized residual|)",
     main = "Scale-location")
lines(lowess(fitted_M1, sqrt(abs(std_M1))), col = "navy", lwd = 2)

## (3) standardized residuals vs the physical driver, Froude
plot(yh$Froude, std_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
     xlab = "Froude", ylab = "standardized residual",
     main = "Standardized residuals vs Froude")
abline(h = c(-2, 0, 2), col = c("grey50", "firebrick", "grey50"), lty = c(3, 1, 3))

## (4) normal Q-Q of the standardized residuals
qqnorm(std_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
       main = "Normal Q-Q (standardized residuals)")
qqline(std_M1, col = "firebrick")
par(op)
```

The diagnostics show that the additive model is **not** homoscedastic on the original scale: the residual spread grows steeply with the fitted value (the scale-location curve slopes sharply upward and the high/low tertile spread ratio printed above is well above 1), and the standardized residuals fan out with `Froude`, neither (task 1 and 3) error model is uniformly correct, so the two are best judged by how well their predictive bands track the data across the whole Froude range.

**Conclusion of part 3.** Under the corrected (additive) error model the hull covariate selected in part 2 is decisively supported: M1 beats M0 by a clear WAIC/PSIS-LOO margin, and $\beta_B$ is credibly negative, a wider beam relative to draught reduces residuary resistance per unit displacement at fixed Froude number. This sharpens part 2's verdict: the weak BAS evidence for `beam_draught` reflected the misspecified homoscedastic log-scale likelihood, not an absence of signal. The Froude-exponent inference, by contrast, is strongly error-model-dependent ($4.69$ vs $\approx 7.2$), which is exactly the heteroscedasticity impact.

## 4. Posterior predictive curves

The last comparison should be made on the original resistance scale, because this is where the data are observed and where the two error assumptions imply visibly different uncertainty bands. For a fair comparison, both bands below are **posterior predictive** bands: each draw includes uncertainty in the mean curve and one new observation-level error term.

For the log-log model, the posterior predictive distribution is lognormal on the original scale: after drawing $(\beta,\sigma^2)$ from the reference-prior posterior, we simulate $$
\log y_i^{\mathrm{rep}} \sim \mathcal N(x_i^\top\beta,\sigma^2),
\qquad y_i^{\mathrm{rep}} = \exp(\log y_i^{\mathrm{rep}}).
$$ For the heteroscedasticity-treated nonlinear model M1, the predictive distribution is additive normal on the original scale: $$
y_i^{\mathrm{rep}} \sim \mathcal N\left(\exp\{b_0 + \gamma_F\widetilde{\log Fn}_i + \beta_B\widetilde{bd}_i\},\sigma^2\right).
$$ The curves hold hull form fixed at a representative value: `beam_draught` is set to its sample median, while the other log-log hull covariates are set to their sample means. This isolates the Froude-resistance curve rather than mixing different hull geometries.

```{r task4-posterior-predictive, fig.width=9, fig.height=5.8}
set.seed(42)

## Froude grid and representative hull values
fr_grid <- seq(min(yh$Froude), max(yh$Froude), length.out = 160)
bd_ref <- median(yh$beam_draught)
hull_ref <- as.list(colMeans(yh[, c("LCB", "prismatic", "L_disp", "beam_draught", "L_beam")]))
hull_ref$beam_draught <- bd_ref

new_loglog <- data.frame(log_Fr = log(fr_grid), hull_ref)
new_M1 <- cbind(intercept = 1,
                log_Fr = log(fr_grid) - mF,
                beam_draught = bd_ref - mB)

## Draw posterior predictive samples for the reference-prior log-log model.
## beta | sigma2, y is normal; sigma2 | y is inverse-gamma.
draw_loglog_pred <- function(fit, newdata, n_draws = 5000) {
  X <- model.matrix(fit)
  y <- model.response(model.frame(fit))
  p <- ncol(X)
  df_res <- fit$df.residual
  beta_hat <- coef(fit)
  s2 <- sum(resid(fit)^2) / df_res
  XtX_inv <- chol2inv(chol(crossprod(X)))
  X_new <- model.matrix(delete.response(terms(fit)), newdata)

  sigma2 <- df_res * s2 / rchisq(n_draws, df = df_res)
  z <- matrix(rnorm(n_draws * p), nrow = n_draws)
  beta <- sweep(z %*% chol(XtX_inv), 1, sqrt(sigma2), `*`)
  beta <- sweep(beta, 2, beta_hat, `+`)

  eta <- beta %*% t(X_new)
  yrep <- exp(eta + matrix(rnorm(length(eta), sd = sqrt(sigma2)),
                          nrow = n_draws, ncol = nrow(X_new)))
  mu <- exp(eta + sigma2 / 2)   # posterior mean of resistance, not median

  list(mean = colMeans(mu),
       lower = apply(yrep, 2, quantile, 0.025),
       upper = apply(yrep, 2, quantile, 0.975))
}

## Draw posterior predictive samples for the additive-error nonlinear M1 model.
draw_M1_pred <- function(fit, X_new, n_draws = 5000) {
  d <- as.matrix(fit)
  d <- d[sample(seq_len(nrow(d)), n_draws), ]
  theta <- d[, paste0("theta[", seq_len(ncol(X_new)), "]"), drop = FALSE]
  sigma <- d[, "sigma"]

  mu <- exp(theta %*% t(X_new))
  yrep <- mu + matrix(rnorm(length(mu), sd = sigma),
                      nrow = n_draws, ncol = nrow(X_new))

  list(mean = colMeans(mu),
       lower = apply(yrep, 2, quantile, 0.025),
       upper = apply(yrep, 2, quantile, 0.975))
}

pp_loglog <- draw_loglog_pred(fit, new_loglog)
pp_M1 <- draw_M1_pred(rw_M1, new_M1)

## Coverage is evaluated at the observed design points, using the same predictive rule.
obs_loglog <- data.frame(log_Fr = yh$log_Fr,
                         LCB = yh$LCB, prismatic = yh$prismatic,
                         L_disp = yh$L_disp, beam_draught = yh$beam_draught,
                         L_beam = yh$L_beam)
obs_M1 <- X1
cover_loglog <- draw_loglog_pred(fit, obs_loglog, n_draws = 3000)
cover_M1 <- draw_M1_pred(rw_M1, obs_M1, n_draws = 3000)
coverage <- rbind(
  `log-log, multiplicative errors` = c(
    coverage = mean(yh$resistance >= cover_loglog$lower & yh$resistance <= cover_loglog$upper),
    mean_band_width = mean(cover_loglog$upper - cover_loglog$lower)),
  `nonlinear M1, additive errors` = c(
    coverage = mean(yh$resistance >= cover_M1$lower & yh$resistance <= cover_M1$upper),
    mean_band_width = mean(cover_M1$upper - cover_M1$lower))
)
round(coverage, 3)

## Plot on the original scale. Negative lower tails from the additive normal
## model are clipped only for display because resistance cannot be negative.
ylim_top <- max(yh$resistance, pp_loglog$upper, pp_M1$upper)
plot(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
     col = rgb(0, 0, 0, 0.42), xlab = "Froude number",
     ylab = "residuary resistance", ylim = c(0, ylim_top * 1.04),
     main = "Posterior predictive curves on the original scale")

col_log <- rgb(0.1, 0.2, 0.7, 0.18)
col_add <- rgb(0.8, 0.1, 0.1, 0.18)
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_loglog$lower, 0), rev(pp_loglog$upper)),
        col = col_log, border = NA)
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_M1$lower, 0), rev(pp_M1$upper)),
        col = col_add, border = NA)
points(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
       col = rgb(0, 0, 0, 0.42))
lines(fr_grid, pp_loglog$mean, col = "navy", lwd = 2.5)
lines(fr_grid, pp_M1$mean, col = "firebrick", lwd = 2.5)
legend("topleft", bty = "n", lwd = c(2.5, 2.5, NA), pch = c(NA, NA, 16),
       pt.cex = c(NA, NA, 0.8),
       col = c("navy", "firebrick", rgb(0, 0, 0, 0.42)),
       legend = c("log-log predictive mean + 95% band",
                  "additive nonlinear M1 predictive mean + 95% band",
                  "observed data"))

pdf("figures/posterior_predictive_curves.pdf", width = 9, height = 5.8)
plot(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
     col = rgb(0, 0, 0, 0.42), xlab = "Froude number",
     ylab = "residuary resistance", ylim = c(0, ylim_top * 1.04),
     main = "Posterior predictive curves on the original scale")
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_loglog$lower, 0), rev(pp_loglog$upper)),
        col = col_log, border = NA)
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_M1$lower, 0), rev(pp_M1$upper)),
        col = col_add, border = NA)
points(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
       col = rgb(0, 0, 0, 0.42))
lines(fr_grid, pp_loglog$mean, col = "navy", lwd = 2.5)
lines(fr_grid, pp_M1$mean, col = "firebrick", lwd = 2.5)
legend("topleft", bty = "n", lwd = c(2.5, 2.5, NA), pch = c(NA, NA, 16),
       pt.cex = c(NA, NA, 0.8),
       col = c("navy", "firebrick", rgb(0, 0, 0, 0.42)),
       legend = c("log-log predictive mean + 95% band",
                  "additive nonlinear M1 predictive mean + 95% band",
                  "observed data"))
dev.off()
```

The plot shows the practical consequence of the two error models. The log-log fit keeps the Froude exponent near the physical fourth-power law and produces multiplicative uncertainty: its band is very tight near zero resistance and expands as resistance grows. This describes the low and middle Froude range well, but its mean curve is too flat at the highest speeds, where the data rise faster than the fitted $Fn^{4.69}$ law.

The additive-error nonlinear fit reacts in the opposite direction. Because it assumes a roughly constant error standard deviation on the original scale, the high-resistance observations receive much more leverage than they had in the log-log likelihood. The fitted curve is therefore steeper ($\gamma_F\approx7.2$) and follows the high-Froude data better. Its predictive band is more nearly constant in absolute width, which is the intended heteroscedasticity correction relative to the log-log model, but this also makes the low-Froude band visibly too wide and can assign some probability to negative resistance before display clipping.

**Conclusion of part 4.** Neither predictive family is perfect across the whole range. The log-log model gives the more natural positive, multiplicative uncertainty at low resistance, while the nonlinear additive model gives a better mean trend at high Froude and makes the selected hull covariate visible. The comparison therefore supports the main conclusion from part 3: inference on the Froude exponent and on `beam_draught` is strongly driven by the error model, so the posterior predictive curves are essential for interpreting the heteroscedasticity correction rather than relying only on coefficient summaries.
