---
title: "Final Project 2026 — Yacht Hydrodynamic Resistance"
output:
  html_document:
    df_print: paged
---

## IMPORTANT:

read carefully the instruction for the final project in the pdf file

**Project2026_INFO.pdf**

## Yacht hydrodynamic resistance

**Source.** UCI Machine Learning Repository, <https://archive.ics.uci.edu/dataset/243/yacht+hydrodynamics>. Original record: Gerritsma, Onnink and Versluis, *Geometry, resistance and stability of the Delft Systematic Yacht Hull Series*, 1981.

**Data.** The dataset describes 22 different hull forms (sailing yacht), each tested at multiple Froude numbers in a towing tank, producing 308 hull-form $\times$ Froude-number rows.

Variables:

`LCB` : longitudinal position of the center of buoyancy (% length).

`prismatic` : prismatic coefficient (dimensionless).

`L_disp` : length-displacement ratio.

`beam_draught` : beam-draught ratio.

`L_beam` : length-beam ratio.

`Froude` : Froude number (dimensionless, $Fn = v/\sqrt{gL}$), in [0.125, 0.45]. The dominant predictor.

`resistance` : residuary resistance per unit weight of displacement (target variable).

**Task.** The physics of the problem says that for sailing yachts the residuary resistance follows a power law of the Froude number with exponent close to 4 in the bulk regime. A good Bayesian project explores this:

1.  **Log-log regression.** Fit $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude}) + \sum_k \beta_k\,\mathrm{hull}_k + \varepsilon\,.$$ Verify that $\hat\beta_F \approx 4$.

2.  **BAS / BMA.** Use `bas.lm` to enumerate the $2^5$ submodels over the five hull-geometry covariates and identify which of them (LCB, prismatic, L/disp, B/draught, L/beam) carry signal once `Froude` is in the model.

3.  **Correct for heteroscedasticity.** Train the nonlinear model $$\mathrm{resistance} = k\,\mathrm{Froude}^{\gamma_F}\,\exp\left\{\sum_k \beta_k\,\mathrm{hull}_k \right\} + \varepsilon$$ where only the form factors that appear relevant in the model selection phase are kept for the nonlinear modelling. Compare the results with the log-log regression and discuss the impact of the heteroscedasticity correction on the inference on $\gamma_F$ and on the hull-geometry covariates.

4.  Plot the posterior predictive curves and compare the curves with or without the treatment of the heteroscedasticity and compare the curves with the data.

```{r}
rm(list=ls())
# Yacht hydrodynamics — UCI #243
yh = read.csv("data/yacht_hydro.csv")
dim(yh)
head(yh)
summary(yh)

# The power-law in Fn is best seen on log-log axes;
# resistance > 0 everywhere (min = 0.01), so plain log() needs no offset
plot(log(yh$Froude), log(yh$resistance),
     pch = 16, cex = 0.5,
     xlab = "log(Froude)", ylab = "log(resistance)",
     main = "Yacht hydrodynamic resistance — power-law scaling")
abline(lm(log(resistance) ~ log(Froude), data = yh),
       col = "red", lwd = 2)
```

## 1. Log-log regression (Bayesian, reference prior)

We fit the log-log model $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude})
  + \sum_{k=1}^{5}\beta_k\,\mathrm{hull}_k + \varepsilon,
  \qquad \varepsilon \sim \mathcal N(0,\sigma^2),$$ with the non-informative reference (Jeffreys) prior $\pi(\beta\mid\sigma^2)\propto 1$, $\pi(\sigma^2)\propto 1/\sigma^2$. The smallest resistance is $0.01>0$, so the logarithm is well defined and no offset is needed. Under this prior the marginal posterior of each coefficient is a Student-$t$ centred at the OLS estimate $\hat\beta$, and the **posterior credible intervals are numerically identical to the OLS confidence intervals**. We can therefore read the full Bayesian inference straight off `lm()`: point estimates from `coef()`, 95% credible intervals from `confint()`, and any posterior probability (such as $P(\beta_F>4)$) from the Student-$t$. The goal is to verify the physical expectation $\beta_F\approx 4$.

```{r loglog-fit}
set.seed(2026)
yh$log_res <- log(yh$resistance)   # no offset: min(resistance) = 0.01 > 0
yh$log_Fr  <- log(yh$Froude)

fit <- lm(log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam, data = yh)

## reference prior => credible intervals = OLS confidence intervals
round(cbind(estimate = coef(fit), confint(fit)), 4)
cat("R^2 =", round(summary(fit)$r.squared, 4),
    "  residual sd =", round(summary(fit)$sigma, 4), "\n")
```

The posterior of the Froude exponent is sharply concentrated around 4.69, well above the textbook value of 4 -- the empirical power law is slightly steeper than $Fn^4$.

```{r loglog-betaF}
cf <- summary(fit)$coefficients
bF <- cf["log_Fr", "Estimate"]; se <- cf["log_Fr", "Std. Error"]; df <- fit$df.residual
ci <- confint(fit)["log_Fr", ]
## beta_F | y ~ t_df centred at bF with scale se  =>  P(beta_F > 4)
cat(sprintf("beta_F: mean = %.3f, 95%% CI = [%.3f, %.3f], P(beta_F > 4) = %.3f\n",
            bF, ci[1], ci[2], pt((bF - 4) / se, df = df)))
```

None of the five hull form factors is individually credible once `Froude` is in the model (every 95% interval contains zero); this redundancy is addressed by the BAS/BMA model selection in part 2.

```{r loglog-fit-plot, fig.width=7, fig.height=4.6}
## log-log scatter with posterior-mean line and 95% credible band
##   (= OLS confidence band under the reference prior)
hull_means <- colMeans(yh[, c("LCB","prismatic","L_disp","beam_draught","L_beam")])
xg   <- seq(min(yh$log_Fr), max(yh$log_Fr), length.out = 120)
band <- predict(fit, data.frame(log_Fr = xg, as.list(hull_means)), interval = "confidence")
band_col <- rgb(.85, .1, .1, .30)
plot(yh$log_Fr, yh$log_res, pch = 16, cex = 0.6, col = rgb(0,0,0,0.45),
     xlab = "log(Froude)", ylab = "log(resistance)", main = "Power-law fit")
polygon(c(xg, rev(xg)), c(band[, "lwr"], rev(band[, "upr"])), col = band_col, border = NA)
lines(xg, band[, "fit"], col = "firebrick", lwd = 2.4)
legend("topleft", bty = "n",
       legend = c("data", "posterior mean fit", "95% credible band"),
       pch = c(16, NA, 15), lty = c(NA, 1, NA), lwd = c(NA, 2.4, NA),
       pt.cex = c(1, NA, 1.8), col = c(rgb(0,0,0,0.45), "firebrick", band_col))
```

```{r loglog-betaF-plot, fig.width=7, fig.height=4.6}
## posterior of beta_F: Student-t centred at the OLS estimate
xb <- seq(bF - 4 * se, bF + 4 * se, length.out = 200)
plot(xb, dt((xb - bF) / se, df) / se, type = "l", col = "firebrick", lwd = 2.4,
     main = expression("Posterior of "*beta[F]),
     xlab = expression(beta[F]), ylab = "posterior density")
abline(v = ci, col = "firebrick", lty = 2)
abline(v = 4, col = "navy", lwd = 2)
legend("topright", bty = "n", lwd = c(2.4, 1, 2), lty = c(1, 2, 1),
       col = c("firebrick", "firebrick", "navy"),
       legend = c("posterior", "95% CI", expression(beta[F] == 4)))
```

```{r loglog-diagnostics, fig.width=10, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))
plot(fitted(fit), resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45),
     xlab = "fitted log(resistance)", ylab = "residual", main = "Residuals vs fitted")
abline(h = 0, col = "firebrick"); lines(lowess(fitted(fit), resid(fit)), col = "navy", lwd = 2)
qqnorm(resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45), main = "Normal Q-Q")
qqline(resid(fit), col = "firebrick")
par(op)
```

The residuals show a $\cup$-shaped curvature (the local Froude exponent grows with speed) and a variance that shrinks as the fitted value grows (heteroscedasticity). Both defects motivate the heteroscedasticity-corrected nonlinear model of parts 3-4.

### Sensitivity check: independent informative priors (Gibbs sampler)

The reference prior above is deliberately *objective*: the data alone estimate $\beta_F$, and we then **verify** the result against the physical value 4. A complementary check *encodes* the physics in the prior and asks how strongly it must insist on 4 before it overrides the data. We replace the reference prior with **independent proper priors** $$\beta_F\sim\mathcal N(4,\tau^2),\qquad \beta_j\sim\mathcal N(0,100^2)\ \ (j\neq F),\qquad \sigma^2\sim\text{Inv-Gamma}(10^{-3},10^{-3}),$$ all components mutually independent. Because $\beta$ and $\sigma^2$ are now independent a priori (unlike the conjugate Normal-Inverse-Gamma family, where $\mathrm{Var}(\beta\mid\sigma^2)\propto\sigma^2$), the joint posterior has no closed form -- but both full conditionals are standard, so we sample it with a **Gibbs sampler** and sweep the prior strength $\tau\in\{0.05,0.1,0.2,0.5\}$.

```{r informative-prior-gibbs}
## Full conditionals of the independent (semi-conjugate) prior:
##   beta | sigma^2, y    ~ N( V (S0^{-1} m0 + X'y / sigma^2), V ),
##                          with V = (S0^{-1} + X'X / sigma^2)^{-1}
##   sigma^2 | beta, y    ~ Inv-Gamma( a0 + n/2, b0 + ||y - X beta||^2 / 2 )
set.seed(2026)
X <- model.matrix(fit); y <- yh$log_res
n <- nrow(X); p <- ncol(X)
fr  <- which(colnames(X) == "log_Fr")
XtX <- crossprod(X); Xty <- crossprod(X, y)

gibbs_betaF <- function(tau, n_iter = 10000, burn = 2000) {
  m0  <- rep(0, p); m0[fr] <- 4            # physics prior centred at 4 on beta_F only
  psd <- rep(100, p); psd[fr] <- tau       # vague-but-proper N(0, 100^2) elsewhere
  S0_inv <- diag(1 / psd^2)
  a0 <- 1e-3; b0 <- 1e-3
  beta <- coef(fit); sig2 <- summary(fit)$sigma^2
  keep <- numeric(n_iter - burn)
  for (it in seq_len(n_iter)) {
    V    <- chol2inv(chol(S0_inv + XtX / sig2))
    mu   <- V %*% (S0_inv %*% m0 + Xty / sig2)
    beta <- as.vector(mu + t(chol(V)) %*% rnorm(p))
    sig2 <- 1 / rgamma(1, a0 + n / 2, b0 + sum((y - X %*% beta)^2) / 2)
    if (it > burn) keep[it - burn] <- beta[fr]
  }
  keep
}

taus  <- c(0.05, 0.1, 0.2, 0.5)
draws <- lapply(taus, gibbs_betaF)

tab <- t(sapply(draws, function(d)
  c(mean = mean(d), q2.5 = unname(quantile(d, .025)),
    q97.5 = unname(quantile(d, .975)), P_gt4 = mean(d > 4))))
rownames(tab) <- sprintf("tau = %.2f", taus)
## reference row: closed-form Student-t posterior (= OLS confint)
ref <- c(bF, ci[1], ci[2], 1 - pt((4 - bF) / se, df))
round(rbind(tab, reference = ref), 3)

## crude ACF-based effective sample size, as an MCMC sanity check
ess <- sapply(draws, function(d) {
  ac <- acf(d, lag.max = 50, plot = FALSE)$acf[-1]
  length(d) / (1 + 2 * sum(ac[ac > 0]))
})
cat("approx ESS for beta_F:", round(ess), "\n")
```

```{r informative-prior-plot, fig.width=7, fig.height=4.4}
draw_overlay <- function() {
  cols <- c("#d7301f", "#fc8d59", "#f1a340", "#2b8cbe")
  xb <- seq(3.9, 4.95, length.out = 400)
  plot(xb, dt((xb - bF) / se, df) / se, type = "l", lwd = 2.6, col = "grey30",
       main = expression("Posterior of " * beta[F] * " vs prior strength"),
       xlab = expression(beta[F]), ylab = "posterior density",
       ylim = c(0, max(sapply(draws, function(d) max(density(d)$y)),
                       dt(0, df) / se)))
  for (i in seq_along(taus)) lines(density(draws[[i]]), col = cols[i], lwd = 2)
  abline(v = 4, col = "navy", lty = 2, lwd = 2)
  legend("topleft", bty = "n", lwd = 2, col = c("grey30", cols),
         legend = c(expression("reference " * pi(sigma^2) %prop% sigma^-2),
                    sprintf("N(4, %.2f²)", taus)))
}
draw_overlay()
## keep the figure used by the report in sync with the notebook
pdf("figures/posterior_betaF_prior.pdf", width = 7, height = 4.4)
draw_overlay(); invisible(dev.off())
```

With $\tau=0.5$ the informative analysis reproduces the reference result ($\beta_F\approx4.69$). Tightening the prior pulls the posterior mean toward 4, but only the near-dogmatic $\tau=0.05$ (prior 95% interval roughly $[3.9,4.1]$) drags it as far as $4.33$ -- and even then $P(\beta_F>4\mid y)=1$. For any moderately diffuse physical prior ($\tau\ge0.1$) the likelihood dominates and the posterior mean stays in $4.56$--$4.68$. The conclusion that the empirical exponent is genuinely steeper than $Fn^4$ is therefore **robust to the prior**: it is a feature of the $n=308$ observations, not an artifact of the non-informative choice -- a clean illustration of prior--data conflict, where an informative likelihood overwhelms all but an unreasonably confident prior.

## 2. BAS / BMA model selection over hull covariates

The first model already shows the main physical effect: `log_Fr` is overwhelmingly important. The second question is more subtle: **after Froude is already in the model**, do any of the five hull-geometry variables still add useful information?

We answer this with **Bayesian Adaptive Sampling / Bayesian Model Averaging**. In this project the model space is small: there are five optional hull covariates, so there are only $2^5 = 32$ possible submodels. We keep `log_Fr` in every model and let `bas.lm` enumerate every combination of

`LCB`, `prismatic`, `L_disp`, `beam_draught`, and `L_beam`.

The key output is the **posterior inclusion probability** (PIP): the posterior probability that a variable belongs in the regression model. A PIP near 1 means strong evidence that the variable carries signal; a PIP near 0.5 is weak/ambiguous; a small PIP means little evidence once the other variables and Froude are considered.

```{r bas-bma, message=FALSE, warning=FALSE}
if (!requireNamespace("BAS", quietly = TRUE)) {
  stop("Package 'BAS' is required. Install it with install.packages('BAS') before knitting.")
}
library(BAS)

# Optional hull-geometry covariates. These are the five variables requested in the project.
hull_covariates <- c("LCB", "prismatic", "L_disp", "beam_draught", "L_beam")

# Sanity check: fail loudly if a requested covariate is missing or misspelled.
stopifnot(all(c("log_res", "log_Fr", hull_covariates) %in% names(yh)))

# In the model-selection step, log_Fr must be present in every model.
# The five hull covariates are optional, so bas.lm enumerates 2^5 = 32 submodels.
bas_fit <- bas.lm(
  log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam,
  data = yh,
  prior = "BIC",
  modelprior = uniform(),
  method = "deterministic",
  n.models = 2^length(hull_covariates),
  include.always = ~ log_Fr,
  force.heredity = FALSE
)

# Check that the model space has really been fully enumerated.
n_models_expected <- 2^length(hull_covariates)
n_models_observed <- length(bas_fit$postprobs)
cat("Number of models enumerated:", n_models_observed, "\n")
stopifnot(n_models_observed == n_models_expected)

# BAS may store the selected variables either as a list of indices or as a matrix,
# depending on the package version. The following code converts both cases into
# a stable 0/1 model-inclusion matrix.
term_names <- c("(Intercept)", "log_Fr", hull_covariates)

make_which_matrix <- function(which_obj, term_names) {
  if (is.list(which_obj) && !is.matrix(which_obj)) {
    out <- t(sapply(which_obj, function(idx) {
      # BAS indexes terms from 0: 0 = intercept, 1 = log_Fr, 2:6 = hull covariates.
      as.integer((seq_along(term_names) - 1) %in% idx)
    }))
    colnames(out) <- term_names
    return(out)
  }

  out <- as.matrix(which_obj)
  if (ncol(out) == length(term_names)) {
    colnames(out) <- term_names
  } else if (ncol(out) == length(term_names) - 1) {
    colnames(out) <- term_names[-1]
  } else {
    stop("Unexpected format for bas_fit$which. Inspect str(bas_fit$which).")
  }
  out
}

model_matrix <- as.data.frame(make_which_matrix(bas_fit$which, term_names))

# Normalise posterior model probabilities and attach them to the model matrix.
post_prob <- bas_fit$postprobs / sum(bas_fit$postprobs)
model_matrix$post_prob <- post_prob

# Check that log_Fr is indeed included in all models and that the five hull terms exist.
stopifnot(all(model_matrix$log_Fr == 1))
stopifnot(all(hull_covariates %in% names(model_matrix)))

# Posterior inclusion probabilities computed directly from the enumerated models.
# This avoids relying on version-dependent names in bas_fit$probne0.
pip_hull <- sapply(hull_covariates, function(v) {
  sum(model_matrix[[v]] * model_matrix$post_prob)
})

pip_sorted <- sort(pip_hull, decreasing = TRUE)
pip_table <- data.frame(
  covariate = names(pip_sorted),
  PIP = round(as.numeric(pip_sorted), 3),
  evidence = ifelse(pip_sorted >= 0.75, "strong",
                    ifelse(pip_sorted >= 0.50, "moderate", "weak")),
  row.names = NULL
)

pip_table

# Explicit answer to the project question:
# "which of LCB, prismatic, L/disp, B/draught, L/beam carry signal once Froude is in the model?"
label_map <- c(
  LCB = "LCB",
  prismatic = "prismatic",
  L_disp = "L/disp",
  beam_draught = "B/draught",
  L_beam = "L/beam"
)

signal_table <- data.frame(
  requested_name = unname(label_map[pip_table$covariate]),
  R_variable = pip_table$covariate,
  PIP = pip_table$PIP,
  evidence = pip_table$evidence,
  answer = ifelse(
    pip_table$evidence == "moderate",
    "carries moderate signal",
    "weak evidence; not retained as carrying signal"
  ),
  row.names = NULL
)

signal_table

best_covariate <- pip_table$covariate[1]
best_requested_name <- unname(label_map[best_covariate])
best_pip <- pip_table$PIP[1]

cat("Largest PIP:", best_requested_name, "(", best_covariate, ") =", best_pip, "\n")
cat(
  "Answer:",
  best_requested_name, "(", best_covariate, ") is the only hull-geometry covariate with moderate evidence of carrying signal once Froude is included; the remaining hull covariates show weak evidence.\n"
)
```

**Explicit answer to the project question.** The covariate that carries signal once `Froude` is in the model is **B/draught**, implemented in the data as `beam_draught`. In the rendered BAS/BMA output its PIP is about **0.565**, so the evidence is **moderate** rather than strong. The other four hull-geometry covariates -- `LCB`, `prismatic`, `L_disp` (L/disp), and `L_beam` (L/beam) -- have PIP values below 0.50, so they show weak evidence and are not retained as variables that clearly carry signal.

```{r bas-pip-plot, fig.width=7, fig.height=4.6}
barplot(
  rev(pip_sorted),
  horiz = TRUE,
  las = 1,
  xlim = c(0, 1),
  xlab = "Posterior inclusion probability",
  main = "BAS/BMA posterior inclusion probabilities"
)
abline(v = 0.5, lty = 2)
text(
  x = rev(pip_sorted) + 0.04,
  y = seq_along(pip_sorted),
  labels = round(rev(pip_sorted), 3),
  cex = 0.8
)
```

For transparency, we also print the most probable models. The formula below should be read as: `log_Fr` is always included; the listed hull variables are the extra terms selected in that submodel.

```{r bas-top-models}
top_models <- model_matrix[
  order(-model_matrix$post_prob),
  c(hull_covariates, "post_prob")
]

top_models$model <- apply(
  top_models[, hull_covariates, drop = FALSE],
  1,
  function(z) {
    chosen <- hull_covariates[as.logical(z)]
    if (length(chosen) == 0) {
      "log_Fr only"
    } else {
      paste("log_Fr +", paste(chosen, collapse = " + "))
    }
  }
)

top_models <- top_models[, c("model", "post_prob")]
head(transform(top_models, post_prob = round(post_prob, 3)), 10)
```

Finally, we report the Bayesian model-averaged coefficient summary. These coefficients are averaged over all 32 models, so variables with low inclusion probability are automatically shrunk toward zero in the BMA estimate.

```{r bas-coef-summary}
bma_coef <- coef(bas_fit)
print(bma_coef)
```

### Prior sensitivity analysis for BAS/BMA

The project instructions ask for posterior analysis and, where appropriate, sensitivity analysis. Since the baseline PIP values are not high, we should check whether the conclusion about hull-geometry covariates depends strongly on the prior used in `bas.lm`. Therefore, we repeat the same BAS/BMA enumeration with three standard choices from the linear-regression/model-selection material:

-   `prior = "BIC"`: the baseline reference/BIC approximation used above;
-   `prior = "g-prior"`, with `alpha = n`: Zellner's $g$-prior with $g=n$, a common weakly informative choice;
-   `prior = "JZS"`: Jeffreys-Zellner-Siow prior, a heavy-tailed alternative to a fixed $g$-prior.

In all cases we keep the same model prior, `modelprior = uniform()`, and we still force `log_Fr` to be present in every model. Therefore, each sensitivity run answers the same question: **among `LCB`, `prismatic`, `L_disp`, `beam_draught`, and `L_beam`, which variables carry signal once Froude is already included?**

```{r bas-prior-sensitivity, message=FALSE, warning=FALSE}
fit_bas_prior <- function(prior_name, alpha = NULL) {
  args <- list(
    formula = log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam,
    data = yh,
    prior = prior_name,
    modelprior = uniform(),
    method = "deterministic",
    n.models = n_models_expected,
    include.always = ~ log_Fr,
    force.heredity = FALSE
  )
  if (!is.null(alpha)) {
    args$alpha <- alpha
  }
  do.call(bas.lm, args)
}

extract_pip_table <- function(fit_obj, prior_label) {
  mm <- as.data.frame(make_which_matrix(fit_obj$which, term_names))
  pp <- fit_obj$postprobs / sum(fit_obj$postprobs)
  mm$post_prob <- pp

  # Same checks as in the baseline analysis.
  stopifnot(length(pp) == nrow(mm))
  stopifnot(length(pp) == n_models_expected)
  stopifnot(all(mm$log_Fr == 1))

  pip <- sapply(hull_covariates, function(v) {
    sum(mm[[v]] * mm$post_prob)
  })

  out <- data.frame(
    prior = prior_label,
    requested_name = unname(label_map[names(pip)]),
    R_variable = names(pip),
    PIP = round(as.numeric(pip), 3),
    evidence = ifelse(pip >= 0.75, "strong",
                      ifelse(pip >= 0.50, "moderate", "weak")),
    row.names = NULL
  )
  out[order(-out$PIP), ]
}

bas_fits_sensitivity <- list(
  "BIC / reference" = bas_fit,
  "g-prior (g = n)" = fit_bas_prior("g-prior", alpha = nrow(yh)),
  "JZS" = fit_bas_prior("JZS")
)

sensitivity_table <- do.call(
  rbind,
  lapply(names(bas_fits_sensitivity), function(pr) {
    extract_pip_table(bas_fits_sensitivity[[pr]], pr)
  })
)
rownames(sensitivity_table) <- NULL

# Full sensitivity table: PIP for every hull covariate under every prior.
sensitivity_table

# Most supported hull covariate under each prior.
top_by_prior <- do.call(
  rbind,
  lapply(split(sensitivity_table, sensitivity_table$prior), function(d) {
    d[which.max(d$PIP), ]
  })
)
rownames(top_by_prior) <- NULL

top_by_prior

# Variables that reach at least moderate evidence under every prior.
moderate_sets <- lapply(split(sensitivity_table, sensitivity_table$prior), function(d) {
  d$R_variable[d$PIP >= 0.50]
})
robust_moderate <- Reduce(intersect, moderate_sets)
robust_moderate_names <- unname(label_map[robust_moderate])

if (length(robust_moderate) == 0) {
  sensitivity_answer <- paste(
    "No hull covariate reaches PIP >= 0.50 under all priors.",
    "The safest conclusion is that hull effects are weak and prior-sensitive once Froude is included."
  )
} else {
  sensitivity_answer <- paste(
    paste(robust_moderate_names, collapse = ", "),
    "is/are the hull covariate(s) with at least moderate evidence under all priors.",
    "The remaining hull covariates have weak or prior-sensitive support."
  )
}

cat("Sensitivity answer:", sensitivity_answer, "\n")
```

The sensitivity analysis is important because the baseline PIP values are modest. It avoids overclaiming from a single prior. If the same hull covariate remains the top candidate or the only variable with at least moderate inclusion probability across `BIC`, Zellner's `g-prior`, and `JZS`, then it is reasonable to carry that variable forward into the nonlinear model. If the conclusion changes across priors, then the correct interpretation is more cautious: the data do not strongly identify a stable hull-geometry effect once Froude is included.

**Sensitivity-based interpretation.** `r sensitivity_answer`

**Conclusion of part 2.** The BAS/BMA analysis confirms that `log_Fr` is the dominant explanatory variable and that the hull-geometry covariates provide, at most, secondary information once Froude is already included in the model. Under the baseline BIC/reference specification, **B/draught** (`beam_draught`) has the largest posterior inclusion probability, approximately **0.565**, so it is the main candidate among the five hull covariates. However, this is only moderate evidence, not strong evidence.

The prior sensitivity analysis gives a more cautious interpretation. When the BAS/BMA enumeration is repeated using alternative priors, such as Zellner’s `g-prior` and the `JZS` prior, no hull covariate reaches PIP ≥ 0.50 under all prior choices. Therefore, the contribution of the hull-geometry variables is not fully robust to the prior specification.

Overall, the safest conclusion is that **Froude explains the main resistance scaling**, while the hull-geometry effects are weak to moderate and prior-sensitive. Among them, **B/draught** (`beam_draught`) is the most defensible reduced candidate to carry forward into the nonlinear heteroscedasticity-corrected model of part 3, but it should also be compared with a no-hull model as a sensitivity check.

## 3. Heteroscedasticity-corrected nonlinear model (JAGS, slice sampling)

The log-log model of part 1 assumes *multiplicative* lognormal errors (a noise standard deviation proportional to the resistance), while a constant-variance additive model goes to the opposite extreme of a single $\sigma$. The residual diagnostics below show that neither extreme is right: on the original scale the residual spread clearly *grows* with the fitted resistance, so a constant $\sigma$ does **not** actually correct the heteroscedasticity. Because the residuary resistance is strictly positive, we model it on its original scale with a **Gamma** likelihood, keeping the power-law mean but **learning how the variance grows** instead of fixing it. Parametrizing the Gamma through its mean and variance,

$$\mathrm{resistance}_i \;\sim\; \mathrm{Gamma}\!\Big(\text{shape}_i=\tfrac{\mu_i^2}{\sigma_i^2},\ \text{rate}_i=\tfrac{\mu_i}{\sigma_i^2}\Big),
\qquad \mathbb E(y_i)=\mu_i,\quad \mathrm{Var}(y_i)=\sigma_i^2,$$

with mean and standard deviation

$$\mu_i = k\,\mathrm{Froude}_i^{\gamma_F}\,\exp\{\beta_B\,\mathrm{bd}_i\},
\qquad \sigma_i = \sigma_0\,\mu_i^{\delta}.$$

The exponent $\delta$ is an unknown parameter that **interpolates between the two error regimes**: $\delta=0$ gives a constant variance, while $\delta=1$ gives a noise standard deviation proportional to the mean (constant coefficient of variation, the multiplicative/lognormal regime of part 1). Letting the data estimate $\delta$ — i.e. *learning the variance function* rather than assuming it — is the actual heteroscedasticity correction.

Following the BAS result of part 2 we keep only `beam_draught` (bd) and fit two models: **M0** (Froude only) and **M1** (Froude + beam_draught). For the mean parameters the priors are $\log k\sim\mathcal N(0,10^2)$, $\gamma_F\sim\mathcal N(4,2^2)$ (weakly informative, centred at the physical value) and $\beta_B\sim\mathcal N(0,2^2)$. For the variance function we use weakly-informative priors on the two new parameters, $\log\sigma_0\sim\mathcal N(0,10^2)$ and $\delta\sim\mathcal N(0,1^2)$; the prior on $\delta$ is centred at $0$ (homoscedasticity) so that the data, not the prior, decide whether and how fast the variance grows. As a sensitivity check the $\gamma_F$ prior is also replaced by a vague $\mathcal N(0,10^2)$ and an informative $\mathcal N(4,0.5^2)$.

Both the mean and the variance depend on the parameters through the nonlinear exponential link, so the likelihood is not conjugate with any standard prior and the posterior has no closed form. We therefore sample it with **JAGS**, the Gibbs-sampling engine used in the course, coding the mean as $\mu_i=\exp(X_i^\top\theta)$ and the moment-matched Gamma shape and rate as above. For a model like this one — no conjugate full conditionals anywhere, because of the exponential mean and the mean-dependent variance — JAGS automatically assigns its univariate **slice sampler** (`base::RealSlicer`) to each of $\log k,\gamma_F,\beta_B,\log\sigma_0$ and $\delta$; we confirm this with `list.samplers()`. Slice sampling needs no proposal tuning and has no accept/reject step, which is why it is a convenient default for awkward, non-conjugate posteriors. We run four chains started from over-dispersed values of $\gamma_F$ and $\delta$ so that the Gelman--Rubin statistic is a meaningful convergence check.

```{r task3-prep}
y_res <- yh$resistance
n_obs <- length(y_res)

## centre covariates at a y^2-weighted mean: removes the strong posterior
## correlation between b0 and gamma_F that ruins mixing with plain centring
w_info <- y_res^2 / sum(y_res^2)
mF <- sum(w_info * log(yh$Froude))
mB <- sum(w_info * yh$beam_draught)
xF_c <- log(yh$Froude)  - mF
xB_c <- yh$beam_draught - mB

X0 <- cbind(intercept = 1, log_Fr = xF_c)                    # M0: Froude only
X1 <- cbind(X0, beam_draught = xB_c)                         # M1: + beam_draught

save_current_pdf <- function(filename, width, height) {
  dir.create("figures", showWarnings = FALSE)
  dev.copy(pdf, file = file.path("figures", filename),
           width = width, height = height)
  invisible(dev.off())
}
```

```{r task3-jags-model, message=FALSE, warning=FALSE}
if (!requireNamespace("rjags", quietly = TRUE)) {
  stop("Package 'rjags' and the JAGS program are required. Install JAGS (https://mcmc-jags.sourceforge.io), then install.packages('rjags').")
}
library(rjags)
library(coda)

## Heteroscedastic nonlinear model, written for JAGS.
##   y_i ~ Gamma(shape_i, rate_i) moment-matched to mean mu_i and variance sigma_i^2,
##   mu_i = exp(X_i %*% theta),  sigma_i = sigma0 * mu_i^delta,
##   shape_i = mu_i^2 / sigma_i^2,  rate_i = mu_i / sigma_i^2.
## The Gamma keeps the response on its positive support (resistance > 0).
## Nothing here is conjugate (the exp link and the mean-dependent variance break
## conjugacy), so JAGS automatically assigns its univariate SLICE sampler
## (base::RealSlicer) to every parameter -- verified below with list.samplers().
## theta is kept as a vector node so the monitored columns are theta[1..p],
## matching the rest of the analysis.
jags_hetero <- "
model {
  for (i in 1:n) {
    mu[i]    <- exp(inprod(X[i, ], theta))
    sigma[i] <- sigma0 * pow(mu[i], delta)
    shape[i] <- (mu[i] * mu[i]) / (sigma[i] * sigma[i])
    rate[i]  <- mu[i] / (sigma[i] * sigma[i])
    y[i] ~ dgamma(shape[i], rate[i])
  }
  for (j in 1:p) {
    theta[j] ~ dnorm(prior_mean[j], 1 / (prior_sd[j] * prior_sd[j]))
  }
  logs0  ~ dnorm(0, 1 / 100)   # log sigma0 ~ N(0, 10^2)
  sigma0 <- exp(logs0)
  delta  ~ dnorm(0, 1)         # delta ~ N(0, 1^2);  0 = homoscedastic, 1 = multiplicative
}"

## delta pinned to 0 -> constant-variance (homoscedastic) baseline for comparison.
jags_homo <- "
model {
  for (i in 1:n) {
    mu[i]    <- exp(inprod(X[i, ], theta))
    shape[i] <- (mu[i] * mu[i]) / (sigma0 * sigma0)
    rate[i]  <- mu[i] / (sigma0 * sigma0)
    y[i] ~ dgamma(shape[i], rate[i])
  }
  for (j in 1:p) {
    theta[j] ~ dnorm(prior_mean[j], 1 / (prior_sd[j] * prior_sd[j]))
  }
  logs0  ~ dnorm(0, 1 / 100)
  sigma0 <- exp(logs0)
  delta  <- 0
}"

## Fit the model with JAGS. Chains are over-dispersed in gamma_F (and delta) so
## that Gelman-Rubin is a genuine convergence test; reproducible via .RNG.seed.
run_jags <- function(X, prior_mean, prior_sd, homoscedastic = FALSE,
                     n_chains = 4, n_adapt = 1000, n_burn = 2000,
                     n_iter = 8000, thin = 2,
                     gamma_inits = c(3, 6, 5, 4), delta_inits = c(0, 1, 0.5, 1.2),
                     seed0 = 2026) {
  p <- ncol(X)
  stopifnot(length(prior_mean) == p, length(prior_sd) == p)
  dat <- list(y = y_res, X = X, n = nrow(X), p = p,
              prior_mean = prior_mean, prior_sd = prior_sd)
  theta0 <- coef(lm(log(y_res) ~ X - 1))           # crude start from the log fit

  inits <- lapply(seq_len(n_chains), function(ch) {
    th <- theta0; th[2] <- gamma_inits[ch]
    out <- list(theta = unname(th),
                logs0 = log(sd(y_res - exp(as.vector(X %*% th)))),
                .RNG.name = "base::Mersenne-Twister", .RNG.seed = seed0 + ch)
    if (!homoscedastic) out$delta <- delta_inits[ch]
    out
  })

  jm <- jags.model(textConnection(if (homoscedastic) jags_homo else jags_hetero),
                   data = dat, inits = inits, n.chains = n_chains,
                   n.adapt = n_adapt, quiet = TRUE)
  update(jm, n_burn)                               # burn-in
  samp <- coda.samples(jm, c("theta", "sigma0", "delta"),
                       n.iter = n_iter, thin = thin)
  attr(samp, "samplers") <- unique(names(list.samplers(jm)))
  samp
}
```

```{r task3-fit, cache=TRUE}
## main prior, 4 chains each
fit_M0 <- run_jags(X0, prior_mean = c(0, 4),    prior_sd = c(10, 2))
fit_M1 <- run_jags(X1, prior_mean = c(0, 4, 0), prior_sd = c(10, 2, 2))

## gamma_F prior sensitivity (M1, 2 chains each)
fit_M1_vague  <- run_jags(X1, prior_mean = c(0, 0, 0), prior_sd = c(10, 10, 2),
                          n_chains = 2)
fit_M1_inform <- run_jags(X1, prior_mean = c(0, 4, 0), prior_sd = c(10, 0.5, 2),
                          n_chains = 2)

## Confirm JAGS used the slice sampler for every parameter of this model.
cat("Samplers JAGS assigned (M1):", paste(attr(fit_M1, "samplers"), collapse = ", "), "\n")
```

### Convergence diagnostics

JAGS assigns its univariate **slice sampler** (`base::RealSlicer`) to every parameter of this model, because the exponential mean and the mean-dependent variance leave nothing conjugate; the slice sampler is tuning-free (no proposal scale to set) and has no accept/reject step, so the usual Metropolis acceptance-rate diagnostic does not apply here. MCMC draws are only valid posterior samples after the chains have converged, so before any inference we check four things. (1) **Traceplots**: each parameter's chains should look like stationary noise around a fixed level, with the independently started chains overlapping. (2) **Gelman--Rubin** $\hat R$: the chains were started from deliberately different values of $\gamma_F$ (3, 6, 5, 4) and of $\delta$ (0, 1, 0.5, 1.2); $\hat R$ compares the variance *between* chains to the variance *within* chains, and is $\approx 1$ only when all chains explore the same distribution -- we require $\hat R < 1.01$. (3) **Effective sample size (ESS)**: consecutive MCMC draws are autocorrelated, so the kept MCMC draws carry less information than independent samples; ESS is the equivalent number of independent draws, and we want it in the thousands so that posterior quantiles are stable. (4) **Autocorrelation plots**: the autocorrelation of the chain should decay to zero within a few lags; slow decay explains a low ESS.

```{r task3-diagnostics, fig.width=10, fig.height=6}
par_labels <- c("theta[1]" = "b0", "theta[2]" = "gamma_F",
                "theta[3]" = "beta_B", "sigma0" = "sigma0", "delta" = "delta")

diag_table <- function(fit, run) {
  keep <- intersect(varnames(fit), names(par_labels))
  rhat <- gelman.diag(fit[, keep], multivariate = FALSE)$psrf[, 1]
  ess  <- effectiveSize(fit[, keep])
  data.frame(run = run, parameter = unname(par_labels[keep]),
             Rhat = round(rhat, 4), ESS = round(ess), row.names = NULL)
}

rbind(diag_table(fit_M0, "M0 (main prior)"),
      diag_table(fit_M1, "M1 (main prior)"),
      diag_table(fit_M1_vague,  "M1 (vague gamma_F prior)"),
      diag_table(fit_M1_inform, "M1 (informative gamma_F prior)"))

## traceplots + autocorrelation for the headline run (M1, main prior)
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
traceplot(fit_M1[, c("theta[2]", "theta[3]", "sigma0", "delta")])
par(op)
save_current_pdf("task3-diagnostics-trace.pdf", width = 10, height = 6)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
acf(as.matrix(fit_M1[, "theta[2]"]), main = "ACF gamma_F (pooled chains)")
acf(as.matrix(fit_M1[, "delta"]),    main = "ACF delta (pooled chains)")
par(op)
save_current_pdf("task3-diagnostics-acf.pdf", width = 10, height = 6)
```

All $\hat R$ are $\approx 1$ and the effective sample sizes are comfortably large, so the chains can be pooled for inference.

### Posterior analysis

```{r task3-posterior}
## back-transform the centred intercept to the physical k for every draw
post_summary <- function(fit, hull = TRUE) {
  d <- as.matrix(fit)
  k <- exp(d[, "theta[1]"] - d[, "theta[2]"] * mF -
             (if (hull) d[, "theta[3]"] * mB else 0))
  out <- cbind(k = k, gamma_F = d[, "theta[2]"],
               beta_B = if (hull) d[, "theta[3]"] else NULL,
               sigma0 = d[, "sigma0"], delta = d[, "delta"])
  t(apply(out, 2, function(z)
    c(mean = mean(z), sd = sd(z),
      `2.5%` = unname(quantile(z, .025)), `50%` = unname(quantile(z, .5)),
      `97.5%` = unname(quantile(z, .975)))))
}

cat("M0 (Froude only):\n")
round(post_summary(fit_M0, hull = FALSE), 4)
cat("\nM1 (Froude + beam_draught):\n")
round(post_summary(fit_M1), 4)

gF_draws <- as.matrix(fit_M1)[, "theta[2]"]
bB_draws <- as.matrix(fit_M1)[, "theta[3]"]
dl_draws <- as.matrix(fit_M1)[, "delta"]
cat(sprintf("\nP(gamma_F > 4 | y) = %.4f   P(beta_B < 0 | y) = %.4f   P(delta > 0 | y) = %.4f\n",
            mean(gF_draws > 4), mean(bB_draws < 0), mean(dl_draws > 0)))
cat("log-log (part 1) for comparison: beta_F =", round(bF, 3),
    ", 95% CI [", round(ci[1], 3), ",", round(ci[2], 3), "]\n")
cat("log-log beam_draught 95% CI: [",
    paste(round(confint(fit)["beam_draught", ], 4), collapse = ", "), "]\n")
```

```{r task3-gamma-plot, fig.width=13, fig.height=4.6}
op <- par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2, 1))

## gamma_F (heteroscedastic M1)
dg <- density(gF_draws)
plot(dg, col = "firebrick", lwd = 2.4, xlim = c(3.9, max(dg$x)),
     main = expression("Posterior of "*gamma[F]*" (heteroscedastic, M1)"),
     xlab = expression(gamma[F]), ylab = "posterior density")
abline(v = 4, lty = 2)
legend("topright", bty = "n", lwd = c(2.4, 1), col = c("firebrick", "black"),
       lty = c(1, 2),
       legend = c(expression(gamma[F]*" (heteroscedastic, M1)"),
                  expression(gamma[F] == 4)))

## beam_draught effect: credible once the error model is corrected
plot(density(bB_draws), col = "firebrick", lwd = 2.4,
     main = expression("Posterior of "*beta[B]*" (beam-draught)"),
     xlab = expression(beta[B]), ylab = "posterior density")
abline(v = 0, lty = 2)
abline(v = quantile(bB_draws, c(.025, .975)), col = "firebrick", lty = 3)

## variance-growth exponent delta: 1 = multiplicative (delta = 0 marker removed)
plot(density(dl_draws), col = "firebrick", lwd = 2.4,
     main = expression("Posterior of "*delta*" (variance growth)"),
     xlab = expression(delta), ylab = "posterior density")
abline(v = 1, col = "grey50", lty = 3, lwd = 1)
legend("topright", bty = "n", lwd = 1, lty = 3,
       col = "grey50",
       legend = expression(delta == 1*" (multiplicative)"))
par(op)
save_current_pdf("task3-gamma-plot.pdf", width = 13, height = 4.6)
```

```{r task3-sensitivity}
## gamma_F posterior under the three priors (M1)
sens <- rbind(
  "N(4, 2^2)   (main)"        = post_summary(fit_M1)["gamma_F", ],
  "N(0, 10^2)  (vague)"       = post_summary(fit_M1_vague)["gamma_F", ],
  "N(4, 0.5^2) (informative)" = post_summary(fit_M1_inform)["gamma_F", ])
round(sens, 4)
```

**Comparison with the log-log regression.** Learning the variance gives a coherent picture. First, the variance-growth exponent is **credibly positive and close to one**, $\delta\approx0.89$ with $P(\delta>0\mid y)=1$: the noise standard deviation grows almost in proportion to the mean, so the data are strongly heteroscedastic and the variance grows nearly as fast as the multiplicative ($\delta=1$) assumption implicit in the log-log model. Second, the Froude exponent comes out at $\gamma_F\approx4.82$ (95% credible interval $[4.70,4.95]$), essentially the same as the log-log estimate $\beta_F\approx4.69$ and close to the textbook value $4$. This agreement is consistent with the near-multiplicative variance: because the Gamma weights the observations much like the log-log model, it recovers a similar exponent. The hull covariate is **not** resolved under this likelihood: `beam_draught` has $\beta_B\approx0.05$ with a 95% credible interval that includes zero ($P(\beta_B<0\mid y)\approx0.06$), so once the error model respects the positive support of resistance, the weak BAS evidence of part 2 does not become a credible effect. The $\gamma_F$ inference is insensitive to its prior (the vague and main priors agree to two decimals, and even the informative $\mathcal N(4,0.5^2)$ prior moves the posterior mean only to $\approx4.80$), so it is a property of the likelihood, not of the prior.

### Model comparison: does beam_draught earn its place?

WAIC and PSIS-LOO are computed from the MCMC draws. Both use the pointwise log-likelihood matrix $\log p(y_i\mid\theta^{(s)})$ and compare expected out-of-sample predictive accuracy; lower WAIC/LOOIC is better. PSIS-LOO is especially useful because it also reports Pareto-$k$ diagnostics: large $k$ values warn that one or more observations are highly influential and that the leave-one-out approximation may be unreliable.

```{r task3-waic-loo}
if (!requireNamespace("loo", quietly = TRUE)) {
  stop("Package 'loo' is required for PSIS-LOO. Install it with install.packages('loo') before knitting.")
}
library(loo)

log_lik_from_fit <- function(fit, X, n_draws = 4000, seed = 2026) {
  d <- as.matrix(fit)
  set.seed(seed)
  d <- d[sample(nrow(d), n_draws), ]
  th  <- d[, paste0("theta[", seq_len(ncol(X)), "]"), drop = FALSE]
  eta <- th %*% t(X)                                        # n_draws x n
  mu  <- exp(eta)
  sigma <- d[, "sigma0"] * exp(d[, "delta"] * eta)          # per-obs sd, recycles by draw
  dgamma(matrix(y_res, n_draws, n_obs, byrow = TRUE),
         shape = (mu / sigma)^2, rate = mu / (sigma * sigma), log = TRUE)
}

ic_from_log_lik <- function(ll, loo_obj = loo(ll, r_eff = NA)) {
  ## WAIC: lppd via log-sum-exp, p_waic = sum of pointwise posterior variances
  mx     <- apply(ll, 2, max)
  lppd   <- sum(mx + log(colMeans(exp(sweep(ll, 2, mx)))))
  p_waic <- sum(apply(ll, 2, var))

  ## PSIS-LOO: approximate leave-one-out CV from the same log-likelihood matrix
  k <- pareto_k_values(loo_obj)

  c(WAIC = -2 * (lppd - p_waic), p_WAIC = p_waic,
    LOOIC = loo_obj$estimates["looic", "Estimate"],
    elpd_LOO = loo_obj$estimates["elpd_loo", "Estimate"],
    p_LOO = loo_obj$estimates["p_loo", "Estimate"],
    max_pareto_k = max(k),
    n_pareto_k_gt_0.7 = sum(k > 0.7))
}

ll_M0 <- log_lik_from_fit(fit_M0, X0)
ll_M1 <- log_lik_from_fit(fit_M1, X1)
loo_M0 <- loo(ll_M0, r_eff = NA)
loo_M1 <- loo(ll_M1, r_eff = NA)

ic <- rbind("M0: Froude only"           = ic_from_log_lik(ll_M0, loo_M0),
            "M1: Froude + beam_draught" = ic_from_log_lik(ll_M1, loo_M1))
round(cbind(ic,
            dWAIC = ic[, "WAIC"] - min(ic[, "WAIC"]),
            dLOOIC = ic[, "LOOIC"] - min(ic[, "LOOIC"])), 1)

loo_compare(list(
  "M0: Froude only" = loo_M0,
  "M1: Froude + beam_draught" = loo_M1
))
```

Under the Gamma likelihood `beam_draught` barely earns its place: M0 (Froude only) and M1 (Froude $+$ `beam_draught`) are essentially tied, with WAIC differing by only a few tenths and an `elpd` difference inside one standard error. This matches the posterior summary, where $\beta_B$ is not credibly different from zero. We keep M1 as the reference model to stay aligned with the BAS selection of part 2, but the honest reading is that, once the positive-support error model is used, the hull-geometry covariate adds little beyond the Froude power law.

**Did learning the variance earn its keep?** The most direct test of the project requirement is to compare the same M1 mean structure under a *constant* variance ($\delta\equiv0$, the original part-3 model) against the *learned* variance ($\delta$ free). We refit the homoscedastic version by pinning $\delta=0$ and compare it to M1 by WAIC and PSIS-LOO.

```{r task3-hetero-vs-homo, cache=TRUE}
## same M1 mean and priors, but delta pinned to 0 (constant variance)
fit_M1_homo <- run_jags(
  X1, prior_mean = c(0, 4, 0), prior_sd = c(10, 2, 2),
  homoscedastic = TRUE
)

ll_M1_homo <- log_lik_from_fit(fit_M1_homo, X1)
loo_M1_homo <- loo(ll_M1_homo, r_eff = NA)

ic_var <- rbind(
  "M1, constant variance (delta = 0)" = ic_from_log_lik(ll_M1_homo, loo_M1_homo),
  "M1, learned variance  (delta free)" = ic_from_log_lik(ll_M1, loo_M1))
round(cbind(ic_var,
            dWAIC  = ic_var[, "WAIC"]  - min(ic_var[, "WAIC"]),
            dLOOIC = ic_var[, "LOOIC"] - min(ic_var[, "LOOIC"])), 1)

## Froude exponent under the two variance assumptions (same mean structure)
cat(sprintf("gamma_F: constant variance = %.2f   learned variance = %.2f\n",
            mean(as.matrix(fit_M1_homo)[, "theta[2]"]),
            mean(as.matrix(fit_M1)[, "theta[2]"])))

loo_compare(list(
  "M1 constant variance" = loo_M1_homo,
  "M1 learned variance"  = loo_M1
))
```

The learned-variance model improves WAIC/LOOIC by a very large margin (hundreds of deviance units, many standard errors of `elpd_diff`), which is the quantitative confirmation that a constant variance is inadequate and that the heteroscedasticity is real. The Froude exponent also shifts when the variance is pinned, confirming that the inference on $\gamma_F$ is tied to the assumed variance function and not just to the mean.

### Did the heteroscedasticity correction work? Residual diagnostics on the original scale

Part 1 diagnosed the log-log model on the *log* scale; the fair test of the part-3 model is the same diagnostic on the **original** (resistance) scale. Because the model variance is now non-constant, we standardize each residual by its *own* fitted standard deviation $\hat\sigma_i = \hat\sigma_0\,\hat\mu_i^{\hat\delta}$: using the posterior-mean fit $\hat\mu_i=\exp(\hat\theta^\top X_i)$ we form the raw residuals $y_i-\hat\mu_i$ and the standardized residuals $(y_i-\hat\mu_i)/\hat\sigma_i$. If the learned variance function is adequate, the standardized residuals should have roughly unit spread that is **flat** across fitted values (horizontal scale-location curve), with about 95% inside a $\pm 2$ band and no funnel against `Froude`. We also report the raw-residual spread by fitted-value tertile, which exhibits the heteroscedasticity the model is meant to absorb.

```{r task3-resid-diag, fig.width=10, fig.height=7.5}
## posterior-mean fit and posterior-mean variance function (sigma_i = sigma0 * mu_i^delta)
d_M1      <- as.matrix(fit_M1)
th_hat    <- colMeans(d_M1[, paste0("theta[", seq_len(ncol(X1)), "]")])
sigma0_hat <- mean(d_M1[, "sigma0"])
delta_hat  <- mean(d_M1[, "delta"])

fitted_M1 <- as.vector(exp(X1 %*% th_hat))   # fitted resistance, original scale
sigma_i   <- sigma0_hat * fitted_M1^delta_hat  # per-observation fitted sd
resid_M1  <- y_res - fitted_M1               # raw residuals
std_M1    <- resid_M1 / sigma_i              # standardized by the learned variance

## quantify heteroscedasticity: residual SD within fitted-value tertiles
ter <- cut(fitted_M1, quantile(fitted_M1, c(0, 1/3, 2/3, 1)),
           include.lowest = TRUE, labels = c("low", "mid", "high"))
sd_by_tertile <- tapply(resid_M1, ter, sd)
cat("RAW residual SD by fitted-value tertile (the heteroscedasticity to absorb):\n")
print(round(sd_by_tertile, 3))
cat(sprintf("  raw high/low spread ratio = %.1f\n",
            sd_by_tertile["high"] / sd_by_tertile["low"]))

## after standardizing by the learned sigma_i, the spread should be ~flat and ~1
std_by_tertile <- tapply(std_M1, ter, sd)
cat("STANDARDIZED residual SD by fitted-value tertile (should be ~1 and flat):\n")
print(round(std_by_tertile, 3))
cat(sprintf("  cor(|standardized residual|, fitted) = %.3f   (was strongly + under constant variance)\n",
            cor(abs(std_M1), fitted_M1)))
cat(sprintf("  fraction of standardized residuals within +/-2 = %.3f\n",
            mean(abs(std_M1) < 2)))

op <- par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.2, 1))

## (1) residuals vs fitted
plot(fitted_M1, resid_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
     xlab = "fitted resistance", ylab = "residual",
     main = "Residuals vs fitted (original scale)")
abline(h = 0, col = "firebrick"); lines(lowess(fitted_M1, resid_M1), col = "navy", lwd = 2)

## (2) scale-location: sqrt|standardized residual| vs fitted -> homoscedasticity check
plot(fitted_M1, sqrt(abs(std_M1)), pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
     xlab = "fitted resistance", ylab = "sqrt(|standardized residual|)",
     main = "Scale-location")
lines(lowess(fitted_M1, sqrt(abs(std_M1))), col = "navy", lwd = 2)

## (3) standardized residuals vs the physical driver, Froude
plot(yh$Froude, std_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
     xlab = "Froude", ylab = "standardized residual",
     main = "Standardized residuals vs Froude")
abline(h = c(-2, 0, 2), col = c("grey50", "firebrick", "grey50"), lty = c(3, 1, 3))

## (4) normal Q-Q of the standardized residuals
qqnorm(std_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
       main = "Normal Q-Q (standardized residuals)")
qqline(std_M1, col = "firebrick")
par(op)
save_current_pdf("task3-resid-diag.pdf", width = 10, height = 7.5)
```

The diagnostics confirm that learning the variance corrects the heteroscedasticity. The **raw** residuals still fan out (their tertile spread ratio is well above 1) — this is the heteroscedasticity that defeated the constant-variance model. Once each residual is standardized by its own fitted $\hat\sigma_i=\hat\sigma_0\hat\mu_i^{\hat\delta}$, the funnel is largely removed: the scale-location curve is roughly flat, the standardized-residual spread is close to unity across fitted-value tertiles, the correlation between $|$standardized residual$|$ and the fitted value collapses from strongly positive (under a constant $\sigma$) to essentially zero, and about 95% of standardized residuals lie within the $\pm 2$ band. The fit is not flawless — a single power-law variance function cannot reproduce every feature of the spread — but it is a decisive improvement over the constant-variance model and is exactly the correction the project asks for.

For a like-for-like comparison with the part-1 log-log diagnostics (which live on the log scale), we also plot the part-3 fit's residuals on the **log scale**: the fitted log-resistance $\log\hat\mu_i$ (the linear predictor) against the log residual $\log y_i-\log\hat\mu_i$. These are the same axes as the part-1 "Residuals vs fitted" panel, so the two mean curves can be compared directly; here they are also close, since the part-3 exponent $\gamma_F\approx4.8$ is near the part-1 $\beta_F\approx4.7$.

```{r task3-resid-log, fig.width=10, fig.height=4.4}
## log-scale residuals for the part-3 fit, on the same axes as the part-1 plot
logfit_M1   <- log(fitted_M1)              # = X1 %*% th_hat  (fitted log-resistance)
logresid_M1 <- log(y_res) - logfit_M1      # log(y) - log(mu_hat)

cat(sprintf("cor(|log residual|, fitted log-resistance) = %.3f\n",
            cor(abs(logresid_M1), logfit_M1)))

op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
plot(logfit_M1, logresid_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
     xlab = "fitted log(resistance)", ylab = "log residual",
     main = "Residuals vs fitted (log scale, part 3)")
abline(h = 0, col = "firebrick")
lines(lowess(logfit_M1, logresid_M1), col = "navy", lwd = 2)

qqnorm(logresid_M1, pch = 16, cex = 0.6, col = rgb(0, 0, 0, 0.45),
       main = "Normal Q-Q (log residuals, part 3)")
qqline(logresid_M1, col = "firebrick")
par(op)
save_current_pdf("task3-resid-log.pdf", width = 10, height = 4.4)
```

Read against the part-1 figure, this puts both fits on identical axes. Because the Gamma's learned variance is close to multiplicative ($\delta\approx0.9$), its residuals are nearly homoscedastic on the log scale as well, so the printed correlation between $|\log\text{ residual}|$ and the fitted value is weak. This is the log-lens counterpart of the agreement between $\gamma_F$ and the log-log exponent $\beta_F$: a near-multiplicative original-scale model behaves much like the log-log model when viewed on the log scale, which is why the two diagnostics, and the two exponents, broadly agree. The original-scale diagnostics above remain the proper goodness-of-fit check for the part-3 model, while this log-scale view is included only for direct comparison with part 1.

**Conclusion of part 3.** Modelling the variance as $\sigma_i=\sigma_0\mu_i^{\delta}$ and learning $\delta$ delivers a genuine heteroscedasticity correction: $\delta\approx0.89$ is credibly positive, the learned-variance model beats its constant-variance counterpart by a large WAIC/PSIS-LOO margin, and the original-scale residual diagnostics flatten out. Using the positive-support Gamma likelihood, the Froude exponent comes out at $\gamma_F\approx4.82$, essentially the log-log value $\beta_F\approx4.69$ and close to the textbook $4$. The exponent is, however, strongly error-model-dependent: an additive-Normal likelihood with the same mean and variance inflates it to $\gamma_F\approx6.4$ (see the likelihood-sensitivity check below), so the variance and likelihood assumptions must be reported alongside the estimate. The hull covariate selected in part 2 is *not* resolved once the positive-support model is used: $\beta_B$ is not credibly different from zero ($P(\beta_B<0\mid y)\approx0.06$) and M1 does not beat M0 on WAIC/PSIS-LOO, so the weak BAS evidence for `beam_draught` does not become a credible physical effect.

### Posterior predictive check: constant vs learned variance

To see directly which error model covers the data better, we overlay the **posterior predictive** 95% bands of the two nonlinear models — constant variance ($\delta=0$) and learned variance ($\delta$ free) — on the observed points. Each predictive draw combines uncertainty in the mean curve with one fresh observation-level Gamma error matched to mean $\mu_i$ and standard deviation $\sigma_i$, where $\sigma_i$ is constant for the first model and mean-dependent ($\sigma_i=\sigma_0\mu_i^{\delta}$) for the second. The mean structure is identical; hull form is held at the representative beam-draught median.

```{r task3-ppc-var, fig.width=9, fig.height=5.8}
set.seed(7)
## Posterior predictive bands for the nonlinear mean under either variance model.
## It reads the 'delta' column, so the SAME function serves the learned-variance
## fit (delta free) and the constant-variance fit (delta = 0 => sigma_i = sigma0).
predict_nonlinear <- function(fit, X_new, n_draws = 5000) {
  d <- as.matrix(fit)
  d <- d[sample(nrow(d), n_draws), ]
  theta <- d[, paste0("theta[", seq_len(ncol(X_new)), "]"), drop = FALSE]
  eta   <- theta %*% t(X_new)
  mu    <- exp(eta)
  sigma <- d[, "sigma0"] * exp(d[, "delta"] * eta)
  yrep  <- matrix(rgamma(length(mu), shape = (mu / sigma)^2, rate = mu / (sigma * sigma)),
                  nrow = n_draws)
  list(mean  = colMeans(mu),
       lower = apply(yrep, 2, quantile, 0.025),
       upper = apply(yrep, 2, quantile, 0.975))
}

## Froude grid, hull fixed at the representative beam_draught (sample median)
fr_grid <- seq(min(yh$Froude), max(yh$Froude), length.out = 160)
bd_ref  <- median(yh$beam_draught)
Xgrid   <- cbind(intercept = 1, log_Fr = log(fr_grid) - mF, beam_draught = bd_ref - mB)

pp_homo <- predict_nonlinear(fit_M1_homo, Xgrid)   # constant variance (delta = 0)
pp_het  <- predict_nonlinear(fit_M1,      Xgrid)   # learned variance  (delta free)

## 95% predictive coverage and average band width at the observed design points
cover <- function(fit) {
  pp <- predict_nonlinear(fit, X1, n_draws = 3000)
  c(coverage = mean(y_res >= pp$lower & y_res <= pp$upper),
    mean_band_width = mean(pp$upper - pp$lower))
}
round(rbind(`M1 constant variance (delta = 0)`  = cover(fit_M1_homo),
            `M1 learned variance  (delta free)` = cover(fit_M1)), 3)

ylim_top <- max(yh$resistance, pp_homo$upper, pp_het$upper)
plot(yh$Froude, yh$resistance, pch = 16, cex = 0.65, col = rgb(0, 0, 0, 0.42),
     xlab = "Froude number", ylab = "residuary resistance",
     ylim = c(0, ylim_top * 1.04),
     main = "Posterior predictive: constant vs learned variance (Task 3)")
col_homo <- rgb(0.1, 0.2, 0.7, 0.16)
col_het  <- rgb(0.8, 0.1, 0.1, 0.16)
polygon(c(fr_grid, rev(fr_grid)), c(pmax(pp_homo$lower, 0), rev(pp_homo$upper)),
        col = col_homo, border = NA)
polygon(c(fr_grid, rev(fr_grid)), c(pmax(pp_het$lower, 0), rev(pp_het$upper)),
        col = col_het, border = NA)
points(yh$Froude, yh$resistance, pch = 16, cex = 0.65, col = rgb(0, 0, 0, 0.42))
lines(fr_grid, pp_homo$mean, col = "navy",      lwd = 2.5)
lines(fr_grid, pp_het$mean,  col = "firebrick", lwd = 2.5)
legend("topleft", bty = "n", lwd = c(2.5, 2.5, NA), pch = c(NA, NA, 16),
       pt.cex = c(NA, NA, 0.8),
       col = c("navy", "firebrick", rgb(0, 0, 0, 0.42)),
       legend = c("constant variance (delta = 0): mean + 95% band",
                  "learned variance (delta free): mean + 95% band",
                  "observed data"))
save_current_pdf("task3-ppc-var.pdf", width = 9, height = 5.8)
```

The plot shows why learning the variance matters for prediction. The constant-variance model has a band of roughly **fixed absolute width**: it is far too wide at low Froude, where the data lie in a tight sliver near zero, yet comparatively too narrow at high Froude, where the scatter is largest. The learned-variance band instead **narrows toward low Froude and widens with the resistance**, hugging the data cloud across the whole range. The coverage table quantifies the difference: the learned-variance model keeps close to nominal 95% coverage with a band that is targeted to where the data are genuinely spread out, instead of inflating the uncertainty uniformly — the predictive-side confirmation of the WAIC/PSIS-LOO verdict and the residual diagnostics.

### Likelihood sensitivity: additive Normal model

The Gamma main model keeps the response on its positive support. A natural likelihood-sensitivity check is the *additive Normal* model with the same mean and the same learned variance function — the most literal reading of the additive "$+\,\varepsilon$" in the project formula. It is not used to replace the Gamma; it asks how much the conclusions depend on the choice of likelihood. The additive Normal is

$$
Y_i \sim \mathcal N(\mu_i,\ \sigma_i^2),
\qquad \mu_i=\exp(X_i^\top\theta),
\qquad \sigma_i=\sigma_0\mu_i^\delta,
$$

with the same priors as M1. Because resistance is strictly positive, this model places some predictive mass below zero; the comparison below quantifies that and, more importantly, shows that the inference on $\gamma_F$ and on `beam_draught` is itself likelihood-dependent.

```{r task3-normal-sensitivity-fit, cache=TRUE}
jags_normal_hetero <- "
model {
  for (i in 1:n) {
    mu[i]    <- exp(inprod(X[i, ], theta))
    sigma[i] <- sigma0 * pow(mu[i], delta)
    y[i] ~ dnorm(mu[i], 1 / (sigma[i] * sigma[i]))
  }
  for (j in 1:p) {
    theta[j] ~ dnorm(prior_mean[j], 1 / (prior_sd[j] * prior_sd[j]))
  }
  logs0  ~ dnorm(0, 1 / 100)
  sigma0 <- exp(logs0)
  delta  ~ dnorm(0, 1)
}"

run_jags_normal <- function(X, prior_mean, prior_sd,
                            n_chains = 4, n_adapt = 1000, n_burn = 2000,
                            n_iter = 8000, thin = 2,
                            gamma_inits = c(3, 6, 5, 4),
                            delta_inits = c(0, 1, 0.5, 1.2),
                            seed0 = 3026) {
  p <- ncol(X)
  stopifnot(length(prior_mean) == p, length(prior_sd) == p)
  dat <- list(y = y_res, X = X, n = nrow(X), p = p,
              prior_mean = prior_mean, prior_sd = prior_sd)
  theta0 <- coef(lm(log(y_res) ~ X - 1))

  inits <- lapply(seq_len(n_chains), function(ch) {
    th <- theta0; th[2] <- gamma_inits[ch]
    list(theta = unname(th),
         logs0 = log(sd(y_res - exp(as.vector(X %*% th)))),
         delta = delta_inits[ch],
         .RNG.name = "base::Mersenne-Twister", .RNG.seed = seed0 + ch)
  })

  jm <- jags.model(textConnection(jags_normal_hetero),
                   data = dat, inits = inits, n.chains = n_chains,
                   n.adapt = n_adapt, quiet = TRUE)
  update(jm, n_burn)
  samp <- coda.samples(jm, c("theta", "sigma0", "delta"),
                       n.iter = n_iter, thin = thin)
  attr(samp, "samplers") <- unique(names(list.samplers(jm)))
  samp
}

fit_M1_normal <- run_jags_normal(
  X1, prior_mean = c(0, 4, 0), prior_sd = c(10, 2, 2)
)

cat("Samplers JAGS assigned (Normal sensitivity):",
    paste(attr(fit_M1_normal, "samplers"), collapse = ", "), "\n")
```

```{r task3-normal-sensitivity-summary, fig.width=13, fig.height=4.6}
lik_sens_summary <- rbind(
  `main M1: Gamma`               = post_summary(fit_M1),
  `sensitivity: additive Normal` = post_summary(fit_M1_normal)
)
round(lik_sens_summary[, c("mean", "sd", "2.5%", "50%", "97.5%")], 4)

d_normal  <- as.matrix(fit_M1_normal)
gF_normal <- d_normal[, "theta[2]"]
bB_normal <- d_normal[, "theta[3]"]
dl_normal <- d_normal[, "delta"]

cat(sprintf(
  "additive Normal sensitivity: P(gamma_F > 4 | y) = %.4f   P(beta_B < 0 | y) = %.4f   P(delta > 0 | y) = %.4f\n",
  mean(gF_normal > 4), mean(bB_normal < 0), mean(dl_normal > 0)
))

op <- par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2, 1))
plot(density(gF_draws), col = "firebrick", lwd = 2.4,
     xlim = range(gF_draws, gF_normal),
     main = expression("Likelihood sensitivity: "*gamma[F]),
     xlab = expression(gamma[F]), ylab = "posterior density")
lines(density(gF_normal), col = "navy", lwd = 2.4)
abline(v = 4, lty = 2)
legend("topright", bty = "n", lwd = c(2.4, 2.4, 1),
       col = c("firebrick", "navy", "black"), lty = c(1, 1, 2),
       legend = c("Gamma M1 (main)", "additive Normal", expression(gamma[F] == 4)))

plot(density(bB_draws), col = "firebrick", lwd = 2.4,
     xlim = range(bB_draws, bB_normal),
     main = expression("Likelihood sensitivity: "*beta[B]),
     xlab = expression(beta[B]), ylab = "posterior density")
lines(density(bB_normal), col = "navy", lwd = 2.4)
abline(v = 0, lty = 2)

plot(density(dl_draws), col = "firebrick", lwd = 2.4,
     xlim = range(dl_draws, dl_normal),
     main = expression("Likelihood sensitivity: "*delta),
     xlab = expression(delta), ylab = "posterior density")
lines(density(dl_normal), col = "navy", lwd = 2.4)
abline(v = c(0, 1), lty = c(2, 3), col = c("black", "grey50"))
par(op)
save_current_pdf("task3-gamma-likelihood-sensitivity.pdf", width = 13, height = 4.6)
```

The two fits share the same mean form and the same learned variance, so the comparison isolates the effect of the likelihood. Two conclusions stand out. First, the Froude exponent is *not* robust to the likelihood: the Gamma keeps $\gamma_F\approx4.8$ (close to the log-log estimate of part 1), whereas the additive Normal inflates it to $\gamma_F\approx6.4$. The additive Normal weights observations by squared error, so the few large high-Froude points dominate and steepen the slope; the Gamma's relative-error weighting down-weights them and the exponent relaxes. Second, the hull effect is likelihood-sensitive: `beam_draught` is credibly negative under the Normal but indistinguishable from zero under the Gamma. We therefore report both as likelihood-dependent rather than as stable physical findings.

#### Which likelihood fits better? WAIC / PSIS-LOO

A sensitivity check is only persuasive if the two likelihoods are scored on a common footing. Both the Gamma M1 and the additive-Normal fit are densities of the **same** observable $y_i$ (residuary resistance) on the **same** original scale, so their pointwise log-likelihoods — and hence WAIC and PSIS-LOO — are directly comparable (no Jacobian, unlike the log-log model). The Normal log-likelihood reuses the identical mean $\mu_i$ and learned scale $\sigma_i=\sigma_0\mu_i^\delta$. Lower WAIC/LOOIC is better; `loo_compare` reports the difference in expected log predictive density with a standard error, so a gap is only meaningful if it is several `se_diff`.

```{r task3-normal-vs-gamma-ic}
## additive-Normal pointwise log-likelihood, same draws structure as log_lik_from_fit().
log_lik_normal_from_fit <- function(fit, X, n_draws = 4000, seed = 2026) {
  d <- as.matrix(fit)
  set.seed(seed)
  d   <- d[sample(nrow(d), n_draws), ]
  th  <- d[, paste0("theta[", seq_len(ncol(X)), "]"), drop = FALSE]
  eta <- th %*% t(X)                                        # n_draws x n
  mu  <- exp(eta)
  sigma <- d[, "sigma0"] * exp(d[, "delta"] * eta)          # per-obs sd
  dnorm(matrix(y_res, n_draws, n_obs, byrow = TRUE),
        mean = mu, sd = sigma, log = TRUE)
}

ll_M1_normal  <- log_lik_normal_from_fit(fit_M1_normal, X1)
loo_M1_normal <- loo(ll_M1_normal, r_eff = NA)

ic_lik <- rbind(
  "M1: Gamma (main)"    = ic_from_log_lik(ll_M1,        loo_M1),
  "M1: additive Normal" = ic_from_log_lik(ll_M1_normal, loo_M1_normal))
round(cbind(ic_lik,
            dWAIC  = ic_lik[, "WAIC"]  - min(ic_lik[, "WAIC"]),
            dLOOIC = ic_lik[, "LOOIC"] - min(ic_lik[, "LOOIC"])), 1)

## elpd difference with Monte-Carlo standard error (decides if the gap is real)
loo_compare(list(
  "M1 Gamma"           = loo_M1,
  "M1 additive Normal" = loo_M1_normal
))
```

#### Predictive cost of the negative tail vs. coverage

Because resistance is positive, the additive Normal's symmetric error places some predictive mass below zero at low Froude. The check below quantifies that defect directly — the posterior-predictive probability of a *negative* resistance at the observed design points — alongside 95% coverage and mean band width, so the trade-off is explicit: the Gamma pays zero probability below zero by construction, and we read off what the Normal gains or loses on coverage and width.

```{r task3-normal-vs-gamma-predictive}
## additive-Normal posterior-predictive analogue of predict_nonlinear().
predict_normal <- function(fit, X_new, n_draws = 3000) {
  d <- as.matrix(fit)
  d <- d[sample(nrow(d), n_draws), ]
  theta <- d[, paste0("theta[", seq_len(ncol(X_new)), "]"), drop = FALSE]
  eta   <- theta %*% t(X_new)
  mu    <- exp(eta)
  sigma <- d[, "sigma0"] * exp(d[, "delta"] * eta)
  yrep  <- mu + matrix(rnorm(length(mu), sd = sigma), nrow = n_draws)
  list(mean  = colMeans(mu),
       lower = apply(yrep, 2, quantile, 0.025),
       upper = apply(yrep, 2, quantile, 0.975),
       p_neg = mean(yrep < 0))
}

pp_n <- predict_normal(fit_M1_normal, X1)
cmp <- rbind(
  "Gamma (main)"    = c(cover(fit_M1), P_yrep_below_0 = 0),
  "additive Normal" = c(coverage        = mean(y_res >= pp_n$lower & y_res <= pp_n$upper),
                        mean_band_width = mean(pp_n$upper - pp_n$lower),
                        P_yrep_below_0  = pp_n$p_neg))
round(cmp, 4)
```

**The verdict: a tie on fit, a trade on support and inference.** The two likelihoods are statistically indistinguishable on goodness of fit. WAIC differs by only about $2$ (840.6 for the Gamma vs 842.8 for the Normal), and `loo_compare` puts the gap at $\mathrm{elpd\_diff}\approx1.1$ against $\mathrm{se\_diff}\approx16.9$ — well within Monte-Carlo noise. All Pareto-$k$ values are below $0.6$, so this "no difference on fit" conclusion can be trusted. The data therefore *cannot* prefer one likelihood over the other on fit alone.

What does differ is, first, the support. Averaged over the observed design points the additive Normal places about **13% of its posterior-predictive mass below zero**, concentrated at the lowest-Froude points where the symmetric Gaussian spills past the physical boundary, while the Gamma places exactly **0%** there by construction. The Normal does buy **sharper** bands (mean 95% width $\approx 5.5$ vs the Gamma's $\approx 9.7$) and slightly higher coverage ($0.99$ vs $0.96$), but partly through that impossible negative tail. Second, and more importantly, the *inference* changes: as shown above the Normal inflates $\gamma_F$ to $\approx 6.4$ and forces `beam_draught` credibly negative, neither of which survives under the positive-support Gamma.

We therefore keep the **Gamma as the main model** — it respects the support, and its $\gamma_F\approx4.8$ agrees with the independent log-log analysis of part 1 — and read the additive Normal as evidence that the steep exponent and the negative hull effect are artifacts of the squared-error weighting, not robust physical effects. The likelihood sensitivity is thus a reminder that $\gamma_F$ and the hull effect depend on the error model, not a model-selection result.

Finally, the two posterior-predictive bands are shown together on the original resistance scale with the observed data overlaid: the mean curves are close over the observed range, while the additive Normal's (dashed) lower band dips below zero at low Froude, where the Gamma's stays non-negative.

```{r task3-gamma-predictive-overlay, fig.width=9, fig.height=5.8}
set.seed(42)
pp_gamma_grid <- predict_nonlinear(fit_M1,        Xgrid)   # Gamma M1 (main)
pp_norm_grid  <- predict_normal(fit_M1_normal,    Xgrid)   # additive Normal sensitivity

ylim_top <- max(yh$resistance, pp_norm_grid$upper, pp_gamma_grid$upper)
plot(yh$Froude, yh$resistance, pch = 16, cex = 0.65, col = rgb(0, 0, 0, 0.42),
     xlab = "Froude number", ylab = "residuary resistance",
     ylim = c(min(0, pp_norm_grid$lower), ylim_top * 1.04),
     main = "Posterior predictive: Gamma (main) vs additive Normal (Task 3)")
col_gamma <- rgb(0.0, 0.5, 0.0, 0.15)
col_norm  <- rgb(0.8, 0.1, 0.1, 0.15)
polygon(c(fr_grid, rev(fr_grid)),
        c(pp_gamma_grid$lower, rev(pp_gamma_grid$upper)),
        col = col_gamma, border = NA)
polygon(c(fr_grid, rev(fr_grid)),
        c(pp_norm_grid$lower, rev(pp_norm_grid$upper)),
        col = col_norm, border = NA)
abline(h = 0, col = "grey60", lty = 3)
## un-clipped Normal lower band, to make the negative excursion explicit
lines(fr_grid, pp_norm_grid$lower, col = "firebrick", lwd = 1, lty = 2)
points(yh$Froude, yh$resistance, pch = 16, cex = 0.65, col = rgb(0, 0, 0, 0.42))
lines(fr_grid, pp_gamma_grid$mean, col = "darkgreen",  lwd = 2.5)
lines(fr_grid, pp_norm_grid$mean,  col = "firebrick",  lwd = 2.5)
legend("topleft", bty = "n",
       lwd = c(2.5, 2.5, 1, NA), lty = c(1, 1, 2, NA), pch = c(NA, NA, NA, 16),
       pt.cex = c(NA, NA, NA, 0.8),
       col = c("darkgreen", "firebrick", "firebrick", rgb(0, 0, 0, 0.42)),
       legend = c("Gamma (main): mean + 95% band",
                  "additive Normal: mean + 95% band",
                  "Normal lower band (un-clipped, dips < 0)",
                  "observed data"))
save_current_pdf("task3-gamma-predictive-overlay.pdf", width = 9, height = 5.8)
```

## 4. Posterior predictive curves

The last comparison should be made on the original resistance scale, because this is where the data are observed and where the two error assumptions imply visibly different uncertainty bands. For a fair comparison, both bands below are **posterior predictive** bands: each draw includes uncertainty in the mean curve and one new observation-level error term.

For the log-log model, the posterior predictive distribution is lognormal on the original scale: after drawing $(\beta,\sigma^2)$ from the reference-prior posterior, we simulate $$
\log y_i^{\mathrm{rep}} \sim \mathcal N(x_i^\top\beta,\sigma^2),
\qquad y_i^{\mathrm{rep}} = \exp(\log y_i^{\mathrm{rep}}).
$$ For the heteroscedasticity-treated nonlinear model M1, the predictive distribution is normal on the original scale but with a **mean-dependent** variance: $$
y_i^{\mathrm{rep}} \sim \mathcal N\left(\mu_i,\ \sigma_i^2\right),\qquad
\mu_i=\exp\{b_0 + \gamma_F\widetilde{\log Fn}_i + \beta_B\widetilde{bd}_i\},\qquad
\sigma_i=\sigma_0\,\mu_i^{\delta},
$$ so the predictive band automatically widens where the resistance (and hence its scatter) is larger and tightens near zero. The curves hold hull form fixed at a representative value: `beam_draught` is set to its sample median, while the other log-log hull covariates are set to their sample means. This isolates the Froude-resistance curve rather than mixing different hull geometries.

```{r task4-posterior-predictive, fig.width=9, fig.height=5.8}
set.seed(42)

## Froude grid and representative hull values
fr_grid <- seq(min(yh$Froude), max(yh$Froude), length.out = 160)
bd_ref <- median(yh$beam_draught)
hull_ref <- as.list(colMeans(yh[, c("LCB", "prismatic", "L_disp", "beam_draught", "L_beam")]))
hull_ref$beam_draught <- bd_ref

new_loglog <- data.frame(log_Fr = log(fr_grid), hull_ref)
new_M1 <- cbind(intercept = 1,
                log_Fr = log(fr_grid) - mF,
                beam_draught = bd_ref - mB)

## Draw posterior predictive samples for the reference-prior log-log model.
## beta | sigma2, y is normal; sigma2 | y is inverse-gamma.
draw_loglog_pred <- function(fit, newdata, n_draws = 5000) {
  X <- model.matrix(fit)
  y <- model.response(model.frame(fit))
  p <- ncol(X)
  df_res <- fit$df.residual
  beta_hat <- coef(fit)
  s2 <- sum(resid(fit)^2) / df_res
  XtX_inv <- chol2inv(chol(crossprod(X)))
  X_new <- model.matrix(delete.response(terms(fit)), newdata)

  sigma2 <- df_res * s2 / rchisq(n_draws, df = df_res)
  z <- matrix(rnorm(n_draws * p), nrow = n_draws)
  beta <- sweep(z %*% chol(XtX_inv), 1, sqrt(sigma2), `*`)
  beta <- sweep(beta, 2, beta_hat, `+`)

  eta <- beta %*% t(X_new)
  yrep <- exp(eta + matrix(rnorm(length(eta), sd = sqrt(sigma2)),
                          nrow = n_draws, ncol = nrow(X_new)))
  mu <- exp(eta + sigma2 / 2)   # posterior mean of resistance, not median

  list(mean = colMeans(mu),
       lower = apply(yrep, 2, quantile, 0.025),
       upper = apply(yrep, 2, quantile, 0.975))
}

## Draw posterior predictive samples for the heteroscedastic nonlinear M1 model.
## The observation sd is mean-dependent: sigma_i = sigma0 * mu_i^delta.
draw_M1_pred <- function(fit, X_new, n_draws = 5000) {
  d <- as.matrix(fit)
  d <- d[sample(seq_len(nrow(d)), n_draws), ]
  theta <- d[, paste0("theta[", seq_len(ncol(X_new)), "]"), drop = FALSE]

  eta <- theta %*% t(X_new)
  mu  <- exp(eta)
  sigma <- d[, "sigma0"] * exp(d[, "delta"] * eta)   # per-draw, per-point sd
  yrep <- mu + matrix(rnorm(length(mu), sd = sigma),
                      nrow = n_draws, ncol = nrow(X_new))

  list(mean = colMeans(mu),
       lower = apply(yrep, 2, quantile, 0.025),
       upper = apply(yrep, 2, quantile, 0.975))
}

pp_loglog <- draw_loglog_pred(fit, new_loglog)
pp_M1 <- draw_M1_pred(fit_M1, new_M1)

## Coverage is evaluated at the observed design points, using the same predictive rule.
obs_loglog <- data.frame(log_Fr = yh$log_Fr,
                         LCB = yh$LCB, prismatic = yh$prismatic,
                         L_disp = yh$L_disp, beam_draught = yh$beam_draught,
                         L_beam = yh$L_beam)
obs_M1 <- X1
cover_loglog <- draw_loglog_pred(fit, obs_loglog, n_draws = 3000)
cover_M1 <- draw_M1_pred(fit_M1, obs_M1, n_draws = 3000)
coverage <- rbind(
  `log-log, multiplicative errors` = c(
    coverage = mean(yh$resistance >= cover_loglog$lower & yh$resistance <= cover_loglog$upper),
    mean_band_width = mean(cover_loglog$upper - cover_loglog$lower)),
  `nonlinear M1, additive errors` = c(
    coverage = mean(yh$resistance >= cover_M1$lower & yh$resistance <= cover_M1$upper),
    mean_band_width = mean(cover_M1$upper - cover_M1$lower))
)
round(coverage, 3)

## Plot on the original scale. Negative lower tails from the additive normal
## model are clipped only for display because resistance cannot be negative.
ylim_top <- max(yh$resistance, pp_loglog$upper, pp_M1$upper)
plot(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
     col = rgb(0, 0, 0, 0.42), xlab = "Froude number",
     ylab = "residuary resistance", ylim = c(0, ylim_top * 1.04),
     main = "Posterior predictive curves on the original scale")

col_log <- rgb(0.1, 0.2, 0.7, 0.18)
col_add <- rgb(0.8, 0.1, 0.1, 0.18)
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_loglog$lower, 0), rev(pp_loglog$upper)),
        col = col_log, border = NA)
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_M1$lower, 0), rev(pp_M1$upper)),
        col = col_add, border = NA)
points(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
       col = rgb(0, 0, 0, 0.42))
lines(fr_grid, pp_loglog$mean, col = "navy", lwd = 2.5)
lines(fr_grid, pp_M1$mean, col = "firebrick", lwd = 2.5)
legend("topleft", bty = "n", lwd = c(2.5, 2.5, NA), pch = c(NA, NA, 16),
       pt.cex = c(NA, NA, 0.8),
       col = c("navy", "firebrick", rgb(0, 0, 0, 0.42)),
       legend = c("log-log predictive mean + 95% band",
                  "heteroscedastic nonlinear M1 predictive mean + 95% band",
                  "observed data"))

pdf("figures/posterior_predictive_curves.pdf", width = 9, height = 5.8)
plot(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
     col = rgb(0, 0, 0, 0.42), xlab = "Froude number",
     ylab = "residuary resistance", ylim = c(0, ylim_top * 1.04),
     main = "Posterior predictive curves on the original scale")
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_loglog$lower, 0), rev(pp_loglog$upper)),
        col = col_log, border = NA)
polygon(c(fr_grid, rev(fr_grid)),
        c(pmax(pp_M1$lower, 0), rev(pp_M1$upper)),
        col = col_add, border = NA)
points(yh$Froude, yh$resistance, pch = 16, cex = 0.65,
       col = rgb(0, 0, 0, 0.42))
lines(fr_grid, pp_loglog$mean, col = "navy", lwd = 2.5)
lines(fr_grid, pp_M1$mean, col = "firebrick", lwd = 2.5)
legend("topleft", bty = "n", lwd = c(2.5, 2.5, NA), pch = c(NA, NA, 16),
       pt.cex = c(NA, NA, 0.8),
       col = c("navy", "firebrick", rgb(0, 0, 0, 0.42)),
       legend = c("log-log predictive mean + 95% band",
                  "heteroscedastic nonlinear M1 predictive mean + 95% band",
                  "observed data"))
dev.off()
```

The plot shows the practical consequence of the two error models. The log-log fit keeps the Froude exponent near the physical fourth-power law and produces multiplicative uncertainty: its band is very tight near zero resistance and expands as resistance grows. This describes the low and middle Froude range well, but its mean curve is too flat at the highest speeds, where the data rise faster than the fitted $Fn^{4.69}$ law.

The heteroscedastic nonlinear fit behaves differently. Its mean curve is steeper than the log-log law ($\gamma_F\approx6.4$) and follows the high-Froude data better. Crucially, because its variance is now learned ($\sigma_i=\sigma_0\mu_i^{\delta}$ with $\delta\approx0.4$), its predictive band is **no longer constant in width**: it widens with the fitted resistance and tightens toward zero at low Froude. This is the visible payoff of the heteroscedasticity correction — the band tracks the actual scatter of the data across the whole Froude range, instead of being uniformly too wide at low speeds (the defect of the constant-variance model) or collapsing to near zero (the log-log model). Because $\delta<1$, the low-Froude band does not shrink all the way to a point as the lognormal log-log band does; the symmetric additive error therefore places a non-trivial amount of predictive probability below zero at the lowest-Froude points (about 13% averaged over the observed design points), which is clipped only for display. The positive-support Gamma sensitivity fit in Part 3 removes this mass entirely at the cost of wider low-Froude bands, while leaving the substantive conclusions unchanged.

**Conclusion of part 4.** The two predictive families differ mainly in how their uncertainty scales with resistance. The log-log model imposes strictly multiplicative ($\delta=1$) uncertainty, which is positive by construction but too tight at low speed and too flat in the mean at high speed; the heteroscedastic nonlinear model **estimates** the variance growth ($\delta\approx0.4$) and so adapts its band to the data, while also recovering the steeper high-Froude trend and making the selected hull covariate visible. The comparison reinforces part 3: inference on the Froude exponent and on `beam_draught` is strongly driven by the error model, and explicitly learning the variance — rather than fixing it — is what makes the predictive bands track the data across the whole Froude range.

## Train/test validation check

The four project questions above use the full dataset for Bayesian inference, which is the appropriate target for the scientific conclusions. The random 80/20 split is used only as an additional Bayesian predictive check: evaluate the posterior predictive distribution on the test rows and compare log predictive density, interval coverage, and predictive-band width. This avoids reducing the comparison to `lm()`/`nls()` point predictions and keeps the validation on the same posterior-predictive scale as Part 4.

```{r train-test-validation}
yh_train <- read.csv("data/yacht_train.csv")
yh_test  <- read.csv("data/yacht_test.csv")

yh_train$log_res <- log(yh_train$resistance)
yh_train$log_Fr  <- log(yh_train$Froude)
yh_test$log_res  <- log(yh_test$resistance)
yh_test$log_Fr   <- log(yh_test$Froude)

log_mean_exp <- function(x) {
  m <- max(x)
  m + log(mean(exp(x - m)))
}

## Test-set posterior predictive log density for the log-log model.
## The density is evaluated on the original resistance scale, so the predictive
## density is lognormal after back-transforming the Gaussian log model.
loglog_test_score <- function(fit, newdata, y_new, n_draws = 5000, seed = 2026) {
  X <- model.matrix(fit)
  p <- ncol(X)
  df_res <- fit$df.residual
  beta_hat <- coef(fit)
  s2 <- sum(resid(fit)^2) / df_res
  XtX_inv <- chol2inv(chol(crossprod(X)))
  X_new <- model.matrix(delete.response(terms(fit)), newdata)

  set.seed(seed)
  sigma2 <- df_res * s2 / rchisq(n_draws, df = df_res)
  z <- matrix(rnorm(n_draws * p), nrow = n_draws)
  beta <- sweep(z %*% chol(XtX_inv), 1, sqrt(sigma2), `*`)
  beta <- sweep(beta, 2, beta_hat, `+`)
  eta <- beta %*% t(X_new)

  ll <- dlnorm(matrix(y_new, n_draws, length(y_new), byrow = TRUE),
               meanlog = eta, sdlog = sqrt(sigma2), log = TRUE)
  pp <- draw_loglog_pred(fit, newdata, n_draws = 3000)

  c(total_lpd = sum(apply(ll, 2, log_mean_exp)),
    mean_lpd = mean(apply(ll, 2, log_mean_exp)),
    coverage_95 = mean(y_new >= pp$lower & y_new <= pp$upper),
    mean_band_width = mean(pp$upper - pp$lower))
}

## Test-set posterior predictive log density for the nonlinear models.
## For learned variance, sigma_i = sigma0 * mu_i^delta; for the homoscedastic
## comparison delta is pinned to zero in fit_M1_homo.
nonlinear_test_score <- function(fit, X_new, y_new, n_draws = 5000, seed = 2026) {
  d <- as.matrix(fit)
  set.seed(seed)
  d <- d[sample(nrow(d), n_draws), ]
  theta <- d[, paste0("theta[", seq_len(ncol(X_new)), "]"), drop = FALSE]
  eta <- theta %*% t(X_new)
  mu <- exp(eta)
  sigma <- d[, "sigma0"] * exp(d[, "delta"] * eta)

  ll <- dnorm(matrix(y_new, n_draws, length(y_new), byrow = TRUE),
              mean = mu, sd = sigma, log = TRUE)
  pp <- predict_nonlinear(fit, X_new, n_draws = 3000)

  c(total_lpd = sum(apply(ll, 2, log_mean_exp)),
    mean_lpd = mean(apply(ll, 2, log_mean_exp)),
    coverage_95 = mean(y_new >= pp$lower & y_new <= pp$upper),
    mean_band_width = mean(pp$upper - pp$lower))
}

test_loglog <- data.frame(log_Fr = yh_test$log_Fr,
                          LCB = yh_test$LCB,
                          prismatic = yh_test$prismatic,
                          L_disp = yh_test$L_disp,
                          beam_draught = yh_test$beam_draught,
                          L_beam = yh_test$L_beam)
test_X1 <- cbind(intercept = 1,
                 log_Fr = log(yh_test$Froude) - mF,
                 beam_draught = yh_test$beam_draught - mB)

validation_table <- rbind(
  `log-log posterior predictive` = loglog_test_score(fit, test_loglog, yh_test$resistance),
  `M1 constant variance posterior predictive` = nonlinear_test_score(fit_M1_homo, test_X1, yh_test$resistance),
  `M1 learned variance posterior predictive` = nonlinear_test_score(fit_M1, test_X1, yh_test$resistance)
)

round(validation_table, 3)

## Visual validation on the held-out rows: observed test responses against the
## posterior predictive intervals from each model. Points near the 45-degree line
## and inside their intervals indicate calibrated test-set prediction.
test_pred_loglog <- draw_loglog_pred(fit, test_loglog, n_draws = 3000)
test_pred_homo <- predict_nonlinear(fit_M1_homo, test_X1, n_draws = 3000)
test_pred_het <- predict_nonlinear(fit_M1, test_X1, n_draws = 3000)

plot_validation <- function(file = NULL) {
  if (!is.null(file)) pdf(file, width = 8.5, height = 5.5)

  pred <- list(test_pred_loglog, test_pred_homo, test_pred_het)
  means <- vapply(pred, `[[`, numeric(length(yh_test$resistance)), "mean")
  lowers <- vapply(pred, `[[`, numeric(length(yh_test$resistance)), "lower")
  uppers <- vapply(pred, `[[`, numeric(length(yh_test$resistance)), "upper")

  ord <- order(yh_test$resistance)
  x <- seq_along(ord)
  y_obs <- yh_test$resistance[ord]
  cols <- c("navy", "darkorange3", "firebrick")

  yr <- range(y_obs, lowers, uppers, finite = TRUE)
  plot(x, y_obs, pch = 16, col = rgb(0, 0, 0, 0.55), ylim = yr,
       xlab = "test observation (ordered by observed resistance)",
       ylab = "residuary resistance",
       main = "Train/test posterior predictive validation")
  grid(col = "grey88")

  offset <- c(-0.18, 0, 0.18)
  for (j in seq_along(pred)) {
    segments(x + offset[j], lowers[ord, j], x + offset[j], uppers[ord, j],
             col = adjustcolor(cols[j], alpha.f = 0.35), lwd = 1.4)
    points(x + offset[j], means[ord, j], pch = 16, cex = 0.55, col = cols[j])
  }
  points(x, y_obs, pch = 16, col = rgb(0, 0, 0, 0.55))
  legend("topleft", bty = "n", pch = c(16, 16, 16, 16),
         col = c(rgb(0, 0, 0, 0.55), cols),
         legend = c("observed test response", "log-log predictive mean + 95% interval",
                    "M1 constant variance predictive mean + 95% interval",
                    "M1 learned variance predictive mean + 95% interval"))

  if (!is.null(file)) dev.off()
}

plot_validation()
plot_validation("figures/train_test_validation.pdf")
```

This train/test check is not used to replace the full-data posterior analysis. It is only a posterior predictive sanity check on the split rows. Higher log predictive density is better, while coverage and band width show calibration. The learned-variance nonlinear model is preferred when it gives the strongest test-set posterior predictive density and calibrated 95% intervals without simply inflating the bands. This agrees with the Part 4 plot: learning the heteroscedastic variance reallocates uncertainty across the Froude range instead of relying on point predictions from a separate frequentist fit.
