---
title: "Final Project 2026 — Yacht Hydrodynamic Resistance"
output:
  html_document:
    df_print: paged
---

## IMPORTANT:

read carefully the instruction for the final project in the pdf file

**Project2026_INFO.pdf**

## Yacht hydrodynamic resistance

**Source.** UCI Machine Learning Repository, <https://archive.ics.uci.edu/dataset/243/yacht+hydrodynamics>. Original record: Gerritsma, Onnink and Versluis, *Geometry, resistance and stability of the Delft Systematic Yacht Hull Series*, 1981.

**Data.** The dataset describes 22 different hull forms (sailing yacht), each tested at multiple Froude numbers in a towing tank, producing 308 hull-form $\times$ Froude-number rows.

Variables:

`LCB` : longitudinal position of the center of buoyancy (% length).

`prismatic` : prismatic coefficient (dimensionless).

`L_disp` : length-displacement ratio.

`beam_draught` : beam-draught ratio.

`L_beam` : length-beam ratio.

`Froude` : Froude number (dimensionless, $Fn = v/\sqrt{gL}$), in [0.125, 0.45]. The dominant predictor.

`resistance` : residuary resistance per unit weight of displacement (target variable).

**Task.** The physics of the problem says that for sailing yachts the residuary resistance follows a power law of the Froude number with exponent close to 4 in the bulk regime. A good Bayesian project explores this:

1.  **Log-log regression.** Fit $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude}) + \sum_k \beta_k\,\mathrm{hull}_k + \varepsilon\,.$$ Verify that $\hat\beta_F \approx 4$.

2.  **BAS / BMA.** Use `bas.lm` to enumerate the $2^5$ submodels over the five hull-geometry covariates and identify which of them (LCB, prismatic, L/disp, B/draught, L/beam) carry signal once `Froude` is in the model.

3.  **Correct for heteroscedasticity.** Train the nonlinear model $$\mathrm{resistance} = k\,\mathrm{Froude}^{\gamma_F}\,\exp\left\{\sum_k \beta_k\,\mathrm{hull}_k \right\} + \varepsilon$$ where only the form factors that appear relevant in the model selection phase are kept for the nonlinear modelling. Compare the results with the log-log regression and discuss the impact of the heteroscedasticity correction on the inference on $\gamma_F$ and on the hull-geometry covariates.

4.  Plot the posterior predictive curves and compare the curves with or without the treatment of the heteroscedasticity and compare the curves with the data.

```{r}
rm(list=ls())
# Yacht hydrodynamics — UCI #243
yh = read.csv("data/yacht_hydro.csv")
dim(yh)
head(yh)
summary(yh)

# The power-law in Fn is best seen on log-log axes;
# resistance > 0 everywhere (min = 0.01), so plain log() needs no offset
plot(log(yh$Froude), log(yh$resistance),
     pch = 16, cex = 0.5,
     xlab = "log(Froude)", ylab = "log(resistance)",
     main = "Yacht hydrodynamic resistance — power-law scaling")
abline(lm(log(resistance) ~ log(Froude), data = yh),
       col = "red", lwd = 2)
```

## 1. Log-log regression (Bayesian, reference prior)

We fit the log-log model $$\log(\mathrm{resistance}) = \beta_0 + \beta_F\,\log(\mathrm{Froude})
  + \sum_{k=1}^{5}\beta_k\,\mathrm{hull}_k + \varepsilon,
  \qquad \varepsilon \sim \mathcal N(0,\sigma^2),$$ with the non-informative reference (Jeffreys) prior $\pi(\beta\mid\sigma^2)\propto 1$, $\pi(\sigma^2)\propto 1/\sigma^2$. The smallest resistance is $0.01>0$, so the logarithm is well defined and no offset is needed. Under this prior the marginal posterior of each coefficient is a Student-$t$ centred at the OLS estimate $\hat\beta$, and the **posterior credible intervals are numerically identical to the OLS confidence intervals**. We can therefore read the full Bayesian inference straight off `lm()`: point estimates from `coef()`, 95% credible intervals from `confint()`, and any posterior probability (such as $P(\beta_F>4)$) from the Student-$t$. The goal is to verify the physical expectation $\beta_F\approx 4$.

```{r loglog-fit}
set.seed(2026)
yh$log_res <- log(yh$resistance)   # no offset: min(resistance) = 0.01 > 0
yh$log_Fr  <- log(yh$Froude)

fit <- lm(log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam, data = yh)

## reference prior => credible intervals = OLS confidence intervals
round(cbind(estimate = coef(fit), confint(fit)), 4)
cat("R^2 =", round(summary(fit)$r.squared, 4),
    "  residual sd =", round(summary(fit)$sigma, 4), "\n")
```

The posterior of the Froude exponent is sharply concentrated around 4.69, well above the textbook value of 4 -- the empirical power law is slightly steeper than $Fn^4$.

```{r loglog-betaF}
cf <- summary(fit)$coefficients
bF <- cf["log_Fr", "Estimate"]; se <- cf["log_Fr", "Std. Error"]; df <- fit$df.residual
ci <- confint(fit)["log_Fr", ]
## beta_F | y ~ t_df centred at bF with scale se  =>  P(beta_F > 4)
cat(sprintf("beta_F: mean = %.3f, 95%% CI = [%.3f, %.3f], P(beta_F > 4) = %.3f\n",
            bF, ci[1], ci[2], pt((bF - 4) / se, df = df)))
```

None of the five hull form factors is individually credible once `Froude` is in the model (every 95% interval contains zero); this redundancy is addressed by the BAS/BMA model selection in part 2.

```{r loglog-fit-plot, fig.width=7, fig.height=4.6}
## log-log scatter with posterior-mean line and 95% credible band
##   (= OLS confidence band under the reference prior)
hull_means <- colMeans(yh[, c("LCB","prismatic","L_disp","beam_draught","L_beam")])
xg   <- seq(min(yh$log_Fr), max(yh$log_Fr), length.out = 120)
band <- predict(fit, data.frame(log_Fr = xg, as.list(hull_means)), interval = "confidence")
band_col <- rgb(.85, .1, .1, .30)
plot(yh$log_Fr, yh$log_res, pch = 16, cex = 0.6, col = rgb(0,0,0,0.45),
     xlab = "log(Froude)", ylab = "log(resistance)", main = "Power-law fit")
polygon(c(xg, rev(xg)), c(band[, "lwr"], rev(band[, "upr"])), col = band_col, border = NA)
lines(xg, band[, "fit"], col = "firebrick", lwd = 2.4)
legend("topleft", bty = "n",
       legend = c("data", "posterior mean fit", "95% credible band"),
       pch = c(16, NA, 15), lty = c(NA, 1, NA), lwd = c(NA, 2.4, NA),
       pt.cex = c(1, NA, 1.8), col = c(rgb(0,0,0,0.45), "firebrick", band_col))
```

```{r loglog-betaF-plot, fig.width=7, fig.height=4.6}
## posterior of beta_F: Student-t centred at the OLS estimate
xb <- seq(bF - 4 * se, bF + 4 * se, length.out = 200)
plot(xb, dt((xb - bF) / se, df) / se, type = "l", col = "firebrick", lwd = 2.4,
     main = expression("Posterior of "*beta[F]),
     xlab = expression(beta[F]), ylab = "posterior density")
abline(v = ci, col = "firebrick", lty = 2)
abline(v = 4, col = "navy", lwd = 2)
legend("topright", bty = "n", lwd = c(2.4, 1, 2), lty = c(1, 2, 1),
       col = c("firebrick", "firebrick", "navy"),
       legend = c("posterior", "95% CI", expression(beta[F] == 4)))
```

```{r loglog-diagnostics, fig.width=10, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2, 1))
plot(fitted(fit), resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45),
     xlab = "fitted log(resistance)", ylab = "residual", main = "Residuals vs fitted")
abline(h = 0, col = "firebrick"); lines(lowess(fitted(fit), resid(fit)), col = "navy", lwd = 2)
qqnorm(resid(fit), pch = 16, cex = 0.5, col = rgb(0,0,0,0.45), main = "Normal Q-Q")
qqline(resid(fit), col = "firebrick")
par(op)
```

The residuals show a $\cup$-shaped curvature (the local Froude exponent grows with speed) and a variance that shrinks as the fitted value grows (heteroscedasticity). Both defects motivate the heteroscedasticity-corrected nonlinear model of parts 3-4.

### Sensitivity check: independent informative priors (Gibbs sampler)

The reference prior above is deliberately *objective*: the data alone estimate $\beta_F$, and we then **verify** the result against the physical value 4. A complementary check *encodes* the physics in the prior and asks how strongly it must insist on 4 before it overrides the data. We replace the reference prior with **independent proper priors** $$\beta_F\sim\mathcal N(4,\tau^2),\qquad \beta_j\sim\mathcal N(0,100^2)\ \ (j\neq F),\qquad \sigma^2\sim\text{Inv-Gamma}(10^{-3},10^{-3}),$$ all components mutually independent. Because $\beta$ and $\sigma^2$ are now independent a priori (unlike the conjugate Normal-Inverse-Gamma family, where $\mathrm{Var}(\beta\mid\sigma^2)\propto\sigma^2$), the joint posterior has no closed form -- but both full conditionals are standard, so we sample it with a **Gibbs sampler** and sweep the prior strength $\tau\in\{0.05,0.1,0.2,0.5\}$.

```{r informative-prior-gibbs}
## Full conditionals of the independent (semi-conjugate) prior:
##   beta | sigma^2, y    ~ N( V (S0^{-1} m0 + X'y / sigma^2), V ),
##                          with V = (S0^{-1} + X'X / sigma^2)^{-1}
##   sigma^2 | beta, y    ~ Inv-Gamma( a0 + n/2, b0 + ||y - X beta||^2 / 2 )
set.seed(2026)
X <- model.matrix(fit); y <- yh$log_res
n <- nrow(X); p <- ncol(X)
fr  <- which(colnames(X) == "log_Fr")
XtX <- crossprod(X); Xty <- crossprod(X, y)

gibbs_betaF <- function(tau, n_iter = 10000, burn = 2000) {
  m0  <- rep(0, p); m0[fr] <- 4            # physics prior centred at 4 on beta_F only
  psd <- rep(100, p); psd[fr] <- tau       # vague-but-proper N(0, 100^2) elsewhere
  S0_inv <- diag(1 / psd^2)
  a0 <- 1e-3; b0 <- 1e-3
  beta <- coef(fit); sig2 <- summary(fit)$sigma^2
  keep <- numeric(n_iter - burn)
  for (it in seq_len(n_iter)) {
    V    <- chol2inv(chol(S0_inv + XtX / sig2))
    mu   <- V %*% (S0_inv %*% m0 + Xty / sig2)
    beta <- as.vector(mu + t(chol(V)) %*% rnorm(p))
    sig2 <- 1 / rgamma(1, a0 + n / 2, b0 + sum((y - X %*% beta)^2) / 2)
    if (it > burn) keep[it - burn] <- beta[fr]
  }
  keep
}

taus  <- c(0.05, 0.1, 0.2, 0.5)
draws <- lapply(taus, gibbs_betaF)

tab <- t(sapply(draws, function(d)
  c(mean = mean(d), q2.5 = unname(quantile(d, .025)),
    q97.5 = unname(quantile(d, .975)), P_gt4 = mean(d > 4))))
rownames(tab) <- sprintf("tau = %.2f", taus)
## reference row: closed-form Student-t posterior (= OLS confint)
ref <- c(bF, ci[1], ci[2], 1 - pt((4 - bF) / se, df))
round(rbind(tab, reference = ref), 3)

## crude ACF-based effective sample size, as an MCMC sanity check
ess <- sapply(draws, function(d) {
  ac <- acf(d, lag.max = 50, plot = FALSE)$acf[-1]
  length(d) / (1 + 2 * sum(ac[ac > 0]))
})
cat("approx ESS for beta_F:", round(ess), "\n")
```

```{r informative-prior-plot, fig.width=7, fig.height=4.4}
draw_overlay <- function() {
  cols <- c("#d7301f", "#fc8d59", "#f1a340", "#2b8cbe")
  xb <- seq(3.9, 4.95, length.out = 400)
  plot(xb, dt((xb - bF) / se, df) / se, type = "l", lwd = 2.6, col = "grey30",
       main = expression("Posterior of " * beta[F] * " vs prior strength"),
       xlab = expression(beta[F]), ylab = "posterior density",
       ylim = c(0, max(sapply(draws, function(d) max(density(d)$y)),
                       dt(0, df) / se)))
  for (i in seq_along(taus)) lines(density(draws[[i]]), col = cols[i], lwd = 2)
  abline(v = 4, col = "navy", lty = 2, lwd = 2)
  legend("topleft", bty = "n", lwd = 2, col = c("grey30", cols),
         legend = c(expression("reference " * pi(sigma^2) %prop% sigma^-2),
                    sprintf("N(4, %.2f²)", taus)))
}
draw_overlay()
## keep the figure used by the report in sync with the notebook
pdf("figures/posterior_betaF_prior.pdf", width = 7, height = 4.4)
draw_overlay(); invisible(dev.off())
```

With $\tau=0.5$ the informative analysis reproduces the reference result ($\beta_F\approx4.69$). Tightening the prior pulls the posterior mean toward 4, but only the near-dogmatic $\tau=0.05$ (prior 95% interval roughly $[3.9,4.1]$) drags it as far as $4.33$ -- and even then $P(\beta_F>4\mid y)=1$. For any moderately diffuse physical prior ($\tau\ge0.1$) the likelihood dominates and the posterior mean stays in $4.56$--$4.68$. The conclusion that the empirical exponent is genuinely steeper than $Fn^4$ is therefore **robust to the prior**: it is a feature of the $n=308$ observations, not an artifact of the non-informative choice -- a clean illustration of prior--data conflict, where an informative likelihood overwhelms all but an unreasonably confident prior.

## 2. BAS / BMA model selection over hull covariates

The first model already shows the main physical effect: `log_Fr` is overwhelmingly important. The second question is more subtle: **after Froude is already in the model**, do any of the five hull-geometry variables still add useful information?

We answer this with **Bayesian Adaptive Sampling / Bayesian Model Averaging**. In this project the model space is small: there are five optional hull covariates, so there are only $2^5 = 32$ possible submodels. We keep `log_Fr` in every model and let `bas.lm` enumerate every combination of

`LCB`, `prismatic`, `L_disp`, `beam_draught`, and `L_beam`.

The key output is the **posterior inclusion probability** (PIP): the posterior probability that a variable belongs in the regression model. A PIP near 1 means strong evidence that the variable carries signal; a PIP near 0.5 is weak/ambiguous; a small PIP means little evidence once the other variables and Froude are considered.

```{r bas-bma, message=FALSE, warning=FALSE}
if (!requireNamespace("BAS", quietly = TRUE)) {
  stop("Package 'BAS' is required. Install it with install.packages('BAS') before knitting.")
}
library(BAS)

# Optional hull-geometry covariates. These are the five variables requested in the project.
hull_covariates <- c("LCB", "prismatic", "L_disp", "beam_draught", "L_beam")

# Sanity check: fail loudly if a requested covariate is missing or misspelled.
stopifnot(all(c("log_res", "log_Fr", hull_covariates) %in% names(yh)))

# In the model-selection step, log_Fr must be present in every model.
# The five hull covariates are optional, so bas.lm enumerates 2^5 = 32 submodels.
bas_fit <- bas.lm(
  log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam,
  data = yh,
  prior = "BIC",
  modelprior = uniform(),
  method = "deterministic",
  n.models = 2^length(hull_covariates),
  include.always = ~ log_Fr,
  force.heredity = FALSE
)

# Check that the model space has really been fully enumerated.
n_models_expected <- 2^length(hull_covariates)
n_models_observed <- length(bas_fit$postprobs)
cat("Number of models enumerated:", n_models_observed, "\n")
stopifnot(n_models_observed == n_models_expected)

# BAS may store the selected variables either as a list of indices or as a matrix,
# depending on the package version. The following code converts both cases into
# a stable 0/1 model-inclusion matrix.
term_names <- c("(Intercept)", "log_Fr", hull_covariates)

make_which_matrix <- function(which_obj, term_names) {
  if (is.list(which_obj) && !is.matrix(which_obj)) {
    out <- t(sapply(which_obj, function(idx) {
      # BAS indexes terms from 0: 0 = intercept, 1 = log_Fr, 2:6 = hull covariates.
      as.integer((seq_along(term_names) - 1) %in% idx)
    }))
    colnames(out) <- term_names
    return(out)
  }

  out <- as.matrix(which_obj)
  if (ncol(out) == length(term_names)) {
    colnames(out) <- term_names
  } else if (ncol(out) == length(term_names) - 1) {
    colnames(out) <- term_names[-1]
  } else {
    stop("Unexpected format for bas_fit$which. Inspect str(bas_fit$which).")
  }
  out
}

model_matrix <- as.data.frame(make_which_matrix(bas_fit$which, term_names))

# Normalise posterior model probabilities and attach them to the model matrix.
post_prob <- bas_fit$postprobs / sum(bas_fit$postprobs)
model_matrix$post_prob <- post_prob

# Check that log_Fr is indeed included in all models and that the five hull terms exist.
stopifnot(all(model_matrix$log_Fr == 1))
stopifnot(all(hull_covariates %in% names(model_matrix)))

# Posterior inclusion probabilities computed directly from the enumerated models.
# This avoids relying on version-dependent names in bas_fit$probne0.
pip_hull <- sapply(hull_covariates, function(v) {
  sum(model_matrix[[v]] * model_matrix$post_prob)
})

pip_sorted <- sort(pip_hull, decreasing = TRUE)
pip_table <- data.frame(
  covariate = names(pip_sorted),
  PIP = round(as.numeric(pip_sorted), 3),
  evidence = ifelse(pip_sorted >= 0.75, "strong",
                    ifelse(pip_sorted >= 0.50, "moderate", "weak")),
  row.names = NULL
)

pip_table

# Explicit answer to the project question:
# "which of LCB, prismatic, L/disp, B/draught, L/beam carry signal once Froude is in the model?"
label_map <- c(
  LCB = "LCB",
  prismatic = "prismatic",
  L_disp = "L/disp",
  beam_draught = "B/draught",
  L_beam = "L/beam"
)

signal_table <- data.frame(
  requested_name = unname(label_map[pip_table$covariate]),
  R_variable = pip_table$covariate,
  PIP = pip_table$PIP,
  evidence = pip_table$evidence,
  answer = ifelse(
    pip_table$evidence == "moderate",
    "carries moderate signal",
    "weak evidence; not retained as carrying signal"
  ),
  row.names = NULL
)

signal_table

best_covariate <- pip_table$covariate[1]
best_requested_name <- unname(label_map[best_covariate])
best_pip <- pip_table$PIP[1]

cat("Largest PIP:", best_requested_name, "(", best_covariate, ") =", best_pip, "\n")
cat(
  "Answer:",
  best_requested_name, "(", best_covariate, ") is the only hull-geometry covariate with moderate evidence of carrying signal once Froude is included; the remaining hull covariates show weak evidence.\n"
)
```

**Explicit answer to the project question.** The covariate that carries signal once `Froude` is in the model is **B/draught**, implemented in the data as `beam_draught`. In the rendered BAS/BMA output its PIP is about **0.565**, so the evidence is **moderate** rather than strong. The other four hull-geometry covariates -- `LCB`, `prismatic`, `L_disp` (L/disp), and `L_beam` (L/beam) -- have PIP values below 0.50, so they show weak evidence and are not retained as variables that clearly carry signal.

```{r bas-pip-plot, fig.width=7, fig.height=4.6}
barplot(
  rev(pip_sorted),
  horiz = TRUE,
  las = 1,
  xlim = c(0, 1),
  xlab = "Posterior inclusion probability",
  main = "BAS/BMA posterior inclusion probabilities"
)
abline(v = 0.5, lty = 2)
text(
  x = rev(pip_sorted) + 0.04,
  y = seq_along(pip_sorted),
  labels = round(rev(pip_sorted), 3),
  cex = 0.8
)
```

For transparency, we also print the most probable models. The formula below should be read as: `log_Fr` is always included; the listed hull variables are the extra terms selected in that submodel.

```{r bas-top-models}
top_models <- model_matrix[
  order(-model_matrix$post_prob),
  c(hull_covariates, "post_prob")
]

top_models$model <- apply(
  top_models[, hull_covariates, drop = FALSE],
  1,
  function(z) {
    chosen <- hull_covariates[as.logical(z)]
    if (length(chosen) == 0) {
      "log_Fr only"
    } else {
      paste("log_Fr +", paste(chosen, collapse = " + "))
    }
  }
)

top_models <- top_models[, c("model", "post_prob")]
head(transform(top_models, post_prob = round(post_prob, 3)), 10)
```

Finally, we report the Bayesian model-averaged coefficient summary. These coefficients are averaged over all 32 models, so variables with low inclusion probability are automatically shrunk toward zero in the BMA estimate.

```{r bas-coef-summary}
bma_coef <- coef(bas_fit)
print(bma_coef)
```

### Prior sensitivity analysis for BAS/BMA

The project instructions ask for posterior analysis and, where appropriate, sensitivity analysis. Since the baseline PIP values are not high, we should check whether the conclusion about hull-geometry covariates depends strongly on the prior used in `bas.lm`. Therefore, we repeat the same BAS/BMA enumeration with three standard choices from the linear-regression/model-selection material:

-   `prior = "BIC"`: the baseline reference/BIC approximation used above;
-   `prior = "g-prior"`, with `alpha = n`: Zellner's $g$-prior with $g=n$, a common weakly informative choice;
-   `prior = "JZS"`: Jeffreys-Zellner-Siow prior, a heavy-tailed alternative to a fixed $g$-prior.

In all cases we keep the same model prior, `modelprior = uniform()`, and we still force `log_Fr` to be present in every model. Therefore, each sensitivity run answers the same question: **among `LCB`, `prismatic`, `L_disp`, `beam_draught`, and `L_beam`, which variables carry signal once Froude is already included?**

```{r bas-prior-sensitivity, message=FALSE, warning=FALSE}
fit_bas_prior <- function(prior_name, alpha = NULL) {
  args <- list(
    formula = log_res ~ log_Fr + LCB + prismatic + L_disp + beam_draught + L_beam,
    data = yh,
    prior = prior_name,
    modelprior = uniform(),
    method = "deterministic",
    n.models = n_models_expected,
    include.always = ~ log_Fr,
    force.heredity = FALSE
  )
  if (!is.null(alpha)) {
    args$alpha <- alpha
  }
  do.call(bas.lm, args)
}

extract_pip_table <- function(fit_obj, prior_label) {
  mm <- as.data.frame(make_which_matrix(fit_obj$which, term_names))
  pp <- fit_obj$postprobs / sum(fit_obj$postprobs)
  mm$post_prob <- pp

  # Same checks as in the baseline analysis.
  stopifnot(length(pp) == nrow(mm))
  stopifnot(length(pp) == n_models_expected)
  stopifnot(all(mm$log_Fr == 1))

  pip <- sapply(hull_covariates, function(v) {
    sum(mm[[v]] * mm$post_prob)
  })

  out <- data.frame(
    prior = prior_label,
    requested_name = unname(label_map[names(pip)]),
    R_variable = names(pip),
    PIP = round(as.numeric(pip), 3),
    evidence = ifelse(pip >= 0.75, "strong",
                      ifelse(pip >= 0.50, "moderate", "weak")),
    row.names = NULL
  )
  out[order(-out$PIP), ]
}

bas_fits_sensitivity <- list(
  "BIC / reference" = bas_fit,
  "g-prior (g = n)" = fit_bas_prior("g-prior", alpha = nrow(yh)),
  "JZS" = fit_bas_prior("JZS")
)

sensitivity_table <- do.call(
  rbind,
  lapply(names(bas_fits_sensitivity), function(pr) {
    extract_pip_table(bas_fits_sensitivity[[pr]], pr)
  })
)
rownames(sensitivity_table) <- NULL

# Full sensitivity table: PIP for every hull covariate under every prior.
sensitivity_table

# Most supported hull covariate under each prior.
top_by_prior <- do.call(
  rbind,
  lapply(split(sensitivity_table, sensitivity_table$prior), function(d) {
    d[which.max(d$PIP), ]
  })
)
rownames(top_by_prior) <- NULL

top_by_prior

# Variables that reach at least moderate evidence under every prior.
moderate_sets <- lapply(split(sensitivity_table, sensitivity_table$prior), function(d) {
  d$R_variable[d$PIP >= 0.50]
})
robust_moderate <- Reduce(intersect, moderate_sets)
robust_moderate_names <- unname(label_map[robust_moderate])

if (length(robust_moderate) == 0) {
  sensitivity_answer <- paste(
    "No hull covariate reaches PIP >= 0.50 under all priors.",
    "The safest conclusion is that hull effects are weak and prior-sensitive once Froude is included."
  )
} else {
  sensitivity_answer <- paste(
    paste(robust_moderate_names, collapse = ", "),
    "is/are the hull covariate(s) with at least moderate evidence under all priors.",
    "The remaining hull covariates have weak or prior-sensitive support."
  )
}

cat("Sensitivity answer:", sensitivity_answer, "\n")
```

The sensitivity analysis is important because the baseline PIP values are modest. It avoids overclaiming from a single prior. If the same hull covariate remains the top candidate or the only variable with at least moderate inclusion probability across `BIC`, Zellner's `g-prior`, and `JZS`, then it is reasonable to carry that variable forward into the nonlinear model. If the conclusion changes across priors, then the correct interpretation is more cautious: the data do not strongly identify a stable hull-geometry effect once Froude is included.

**Sensitivity-based interpretation.** `r sensitivity_answer`

**Conclusion of part 2.** The BAS/BMA analysis confirms that `log_Fr` is the dominant explanatory variable and that the hull-geometry covariates provide, at most, secondary information once Froude is already included in the model. Under the baseline BIC/reference specification, **B/draught** (`beam_draught`) has the largest posterior inclusion probability, approximately **0.565**, so it is the main candidate among the five hull covariates. However, this is only moderate evidence, not strong evidence.

The prior sensitivity analysis gives a more cautious interpretation. When the BAS/BMA enumeration is repeated using alternative priors, such as Zellner’s `g-prior` and the `JZS` prior, no hull covariate reaches PIP ≥ 0.50 under all prior choices. Therefore, the contribution of the hull-geometry variables is not fully robust to the prior specification.

Overall, the safest conclusion is that **Froude explains the main resistance scaling**, while the hull-geometry effects are weak to moderate and prior-sensitive. Among them, **B/draught** (`beam_draught`) is the most defensible reduced candidate to carry forward into the nonlinear heteroscedasticity-corrected model of part 3, but it should also be compared with a no-hull model as a sensitivity check.
